---
title: "SpectraWeaveR"
subtitle: "End-to-End Pipeline for 40-Color Spectral Flow Cytometry Analysis"
---

## Overview

SpectraWeaveR orchestrates five established Bioconductor/GitHub packages into
a single reproducible pipeline for spectral flow cytometry data — from spectral
unmixing through clustering and visualization.

| Step | Function | Package |
|------|----------|---------|
| 1. Unmixing | `sw_unmix_pipeline()`, `sw_unmix()`, `sw_load_unmixed()` | AutoSpectral / flowCore |
| 2. Gating | `sw_gate()`, `sw_build_gating_template()` | openCyto / flowWorkspace |
| 3. Signal QC | `sw_signal_qc()`, `sw_signal_qc_batch()` | PeacoQC |
| 4. Batch correction | `sw_batch_correct()`, `sw_prepare_for_correction()` | cyCombine |
| 5. Clustering | `sw_cluster()`, `sw_plot_clusters()` | FlowSOM |
| End-to-end | `run_pipeline()` | all of the above |

## Quick start

### Full pipeline

```{r}
#| eval: false
library(SpectraWeaveR)

# Define sample metadata
sample_meta <- data.frame(
  file     = c("sample1.fcs", "sample2.fcs", "sample3.fcs", "sample4.fcs"),
  sample   = c("S1", "S2", "S3", "S4"),
  batch    = c("B1", "B1", "B2", "B2"),
  condition = c("Ctrl", "Stim", "Ctrl", "Stim")
)

# Define markers
markers <- c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14", "CD16",
             "CD45RA", "CD45RO", "HLA-DR")
lineage_markers <- c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14")

# Run the full pipeline
results <- run_pipeline(
  fcs_dir         = "path/to/unmixed_fcs/",
  sample_meta     = sample_meta,
  markers         = markers,
  lineage_markers = lineage_markers,
  cofactor        = 6000,
  n_metaclusters  = 20,
  seed            = 42
)

# Access results
results$cluster_assignments  # per-cell metacluster labels
results$cluster_mfis         # MFI per metacluster
results$corrected            # batch-corrected expression data
```

### Step-by-step usage

```{r}
#| eval: false
library(SpectraWeaveR)

# 1. Load unmixed FCS files
fs <- sw_load_unmixed("path/to/unmixed_fcs/")

# 2. Remove margin events (before transformation!)
ff_list <- lapply(seq_along(fs), function(i) sw_remove_margins(fs[[i]]))
names(ff_list) <- flowCore::sampleNames(fs)

# 3. Signal QC
qc_results <- sw_signal_qc_batch(ff_list)
qc_summary <- sw_qc_summary(qc_results, threshold = 30)
ff_clean <- qc_results$cleaned

# 4. Batch correction
uncorrected <- sw_prepare_for_correction(
  ff_clean, sample_meta, markers, cofactor = 6000
)
corrected <- sw_batch_correct(uncorrected, markers, covar = "condition")

# 5. Clustering
fsom <- sw_cluster(corrected, lineage_markers, n_metaclusters = 20)
assignments <- sw_get_cluster_assignments(fsom)
mfis <- sw_cluster_mfis(fsom)
sw_plot_clusters(fsom, "clusters.pdf")
```

## Key design decisions

- **`truncate_max_range = FALSE`** is enforced when reading FCS files — Aurora
  fluorescence intensities (~4×10⁶) exceed flowCore's default range cutoffs.
- **arcsinh cofactor = 6000** is used for spectral flow cytometry
  (not 5 as in CyTOF).
- Format conversion utilities (`sw_flowframe_to_tibble()`,
  `sw_tibble_to_flowframe()`) bridge the incompatible data structures required
  by each tool.

## Learn more

- [Installation](installation.qmd) — how to install SpectraWeaveR and its
  dependencies.
- [Spectral Unmixing vignette](vignettes/spectral-unmixing.qmd) — complete
  walkthrough of the unmixing workflow with AutoSpectral.
- [API Reference](reference/index.qmd) — documentation for every exported
  function.
