---
title: "Installation"
---

## Prerequisites

SpectraWeaveR requires **R ≥ 4.1.0** and relies on packages from CRAN,
Bioconductor, and GitHub. Follow the sections below in order.

## 1. Install Bioconductor dependencies

SpectraWeaveR imports several Bioconductor packages. Install them first using
BiocManager:
```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
  "flowCore",       # FCS I/O, flowFrame/flowSet classes
  "flowWorkspace",  # GatingSet, cytoset
  "openCyto",       # CSV-driven hierarchical gating
  "PeacoQC"         # Isolation Tree + MAD signal QC
))

# Install kohonen from CRAN
install.packages("kohonen")  # Self-organizing map clustering
```

## 2. Install GitHub dependencies

[cyCombine](https://github.com/biosurf/cyCombine) provides ComBat-based batch
correction and is only available on GitHub:
```{r}
#| eval: false
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("biosurf/cyCombine")
```

## 3. Install SpectraWeaveR
```{r}
#| eval: false
remotes::install_github("mdmanurung/SpectraWeaveR")
```

## Optional dependencies

### AutoSpectral (spectral unmixing)

If you need spectral unmixing from raw data (Step 1 of the pipeline), install
[AutoSpectral](https://github.com/DrCytometer/AutoSpectral):
```{r}
#| eval: false
# Core unmixing engine
remotes::install_github("DrCytometer/AutoSpectral")

# C++ acceleration (strongly recommended)
remotes::install_github("DrCytometer/AutoSpectralRcpp")
```

### S7 (composable pipeline framework)

The composable pipeline uses [S7](https://rconsortium.github.io/S7/) classes.
Install it if you want to use `sw_pipeline_step()`, `sw_pipeline()`, and related
functions:
```{r}
#| eval: false
install.packages("S7")
```

### Visualization extras

```{r}
#| eval: false
# Gate visualisation
BiocManager::install("ggcyto")

# UMAP dimensionality reduction
install.packages("uwot")

# Cluster annotation heatmaps
install.packages("pheatmap")
```

### FastPG (graph-based clustering)

An alternative clustering backend. Install if you want to use
`sw_cluster_run(..., method = "fastpg")`:
```{r}
#| eval: false
remotes::install_github("sararselitsky/FastPG")
```

## Verify installation

Check that the core package loads without errors:
```{r}
#| eval: false
library(SpectraWeaveR)
packageVersion("SpectraWeaveR")
#> [1] '0.1.0'
```

You can also verify key dependencies:
```{r}
#| eval: false
sapply(c("flowCore", "flowWorkspace", "openCyto",
         "PeacoQC", "kohonen", "cyCombine"),
       packageVersion)
```

## Dependency summary

| Package | Source | Required? | Purpose |
|---------|--------|-----------|---------|
| flowCore ≥ 2.6.0 | Bioconductor | Yes | FCS I/O, flowFrame/flowSet |
| flowWorkspace ≥ 4.6.0 | Bioconductor | Yes | GatingSet, cytoset |
| openCyto ≥ 2.6.0 | Bioconductor | Yes | Automated gating |
| PeacoQC ≥ 1.4.0 | Bioconductor | Yes | Signal quality control |
| kohonen ≥ 3.0.0 | CRAN | Yes | SOM clustering |
| cyCombine ≥ 0.2.7 | GitHub | Yes | Batch correction |
| dplyr ≥ 1.0.0 | CRAN | Yes | Data manipulation |
| tibble ≥ 3.1.0 | CRAN | Yes | Tibble data frames |
| AutoSpectral ≥ 1.5.0 | GitHub | Optional | Spectral unmixing |
| AutoSpectralRcpp ≥ 1.0.0 | GitHub | Optional | C++ acceleration for unmixing |
| FastPG | GitHub | Optional | Graph-based clustering |
| S7 ≥ 0.2.0 | CRAN | Optional | Composable pipeline (R7 OOP) |
| ggcyto ≥ 1.22.0 | Bioconductor | Optional | Gate visualisation |
| ggplot2 | CRAN | Optional | Plotting |
| uwot ≥ 0.1.14 | CRAN | Optional | UMAP |
| pheatmap ≥ 1.0.12 | CRAN | Optional | Heatmaps |
