---
title: "Batch Correction with SpectraWeaveR and cyCombine"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Batch effects are systematic, non-biological differences between samples
processed on different days, instruments, or with different reagent lots.
In high-parameter spectral flow cytometry, batch effects can mask true
biological signals and confound downstream analysis.

SpectraWeaveR provides a complete batch correction workflow powered by
[cyCombine](https://github.com/biosurf/cyCombine), which applies
ComBat batch correction within Self-Organising Map (SOM) clusters.
This cluster-aware approach preserves biological heterogeneity while
removing technical variation.

### Two workflows

SpectraWeaveR supports both an **all-in-one** workflow and a
**modular** workflow for batch correction:

| Workflow | Functions | When to use |
|----------|-----------|-------------|
| All-in-one | `sw_batch_correct()` | Standard batch correction with defaults |
| Modular | `sw_normalize()` → `sw_create_som()` → `sw_correct_data()` | When you need to tune normalization, SOM grid size, or use custom cluster labels |

### The full batch correction function set

| Function | Purpose |
|----------|---------|
| `sw_prepare_for_correction()` | Convert flowFrame list → tibble with arcsinh transformation |
| `sw_batch_correct()` | All-in-one ComBat batch correction |
| `sw_normalize()` | Per-batch normalization (scale/rank) |
| `sw_create_som()` | SOM clustering for ComBat |
| `sw_correct_data()` | ComBat correction with pre-computed labels |
| `sw_detect_batch_effect()` | Diagnostic density + MDS + EMD plots |
| `sw_compute_emd()` | Earth Mover's Distance per marker/cluster |
| `sw_evaluate_emd()` | Compare EMD before vs. after correction |
| `sw_evaluate_mad()` | Compare MAD before vs. after correction |
| `sw_evaluate_correction()` | Quick MAD-based quality check |
| `sw_plot_batch_densities()` | Density ridgeline plots by batch |
| `sw_plot_batch_dimred()` | UMAP/PCA colored by batch |

## Prerequisites

### Install cyCombine

cyCombine is available on GitHub:

```{r}
#| label: install-cycombine
# Ensure BioConductor packages are available
setRepositories(ind = 1:2)
BiocManager::install(c("flowCore", "Biobase", "sva"))

# Install cyCombine
remotes::install_github("biosurf/cyCombine")
```

## Prepare input data

### From FCS files

The typical starting point is a directory of unmixed FCS files (e.g.,
exported from SpectroFlo or produced by `sw_unmix_pipeline()`) along with
a metadata table describing each sample's batch and condition.

```{r}
#| label: prepare-data
library(SpectraWeaveR)

# Load unmixed FCS files
fcs_dir <- "path/to/unmixed_fcs"
fs <- sw_load_unmixed(fcs_dir)

# Define markers of interest
markers <- c(
  "CD3", "CD4", "CD8", "CD19", "CD20", "CD14", "CD56",
  "CD45RA", "CD45RO", "CD27", "CD127", "CD25", "HLA-DR",
  "CD69", "CD279", "CD197", "KLRG1", "GranzymeB"
)

# Create a metadata table
# Each row matches a sample name in the flowSet
sample_meta <- data.frame(
  sample = sampleNames(fs),
  batch = c(rep("Batch1", 5), rep("Batch2", 5)),
  condition = c(rep("Healthy", 3), rep("Disease", 2),
                rep("Healthy", 3), rep("Disease", 2))
)

# Convert flowSet to a list of flowFrames for sw_prepare_for_correction
ff_list <- setNames(
  lapply(sampleNames(fs), function(s) fs[[s]]),
  sampleNames(fs)
)
```

### Prepare the tibble

`sw_prepare_for_correction()` performs three operations:

1. Extracts the specified marker columns from each flowFrame
2. Applies arcsinh transformation with the spectral flow cofactor (6000)
3. Adds metadata columns (`sample`, `batch`, `condition`)

```{r}
#| label: prepare-tibble
uncorrected <- sw_prepare_for_correction(
  ff_list = ff_list,
  sample_meta = sample_meta,
  markers = markers,
  cofactor = 6000  # Spectral flow cytometry default (not 5 as for CyTOF)
)

head(uncorrected)
# Save checkpoint
saveRDS(uncorrected, "uncorrected.rds")
```

### From a pre-existing tibble

If your data is already in tibble format (e.g., from another tool or a
CSV export), ensure it has the required columns and skip straight to
batch correction:
```{r}
#| label: from-tibble
# Your tibble must have:
# - Marker columns (numeric, already transformed)
# - A 'batch' column
# - Optionally a 'sample' and 'condition' column
str(uncorrected)
```

## Detect batch effects

Before correcting, it is good practice to visualise the batch effects to
confirm they exist and understand their magnitude.

```{r}
#| label: detect-batch
# Diagnostic plots: density ridgelines, per-marker EMD, and MDS
diagnostics <- sw_detect_batch_effect(
  df = uncorrected,
  markers = markers,
  out_dir = "batch_diagnostics"  # Saves PNGs; use NULL for ggplot objects
)
```

You can also visualise batch effects with UMAP or PCA:

```{r}
#| label: umap-before
# UMAP colored by batch (uncorrected)
set.seed(42)
p_before <- sw_plot_batch_dimred(
  df = uncorrected[sample(nrow(uncorrected), 20000), ],
  markers = markers,
  type = "umap",
  name = "Uncorrected"
)
p_before
```

## All-in-one batch correction

For a standard correction with sensible defaults, `sw_batch_correct()`
handles normalization, SOM clustering, and ComBat in a single call:

```{r}
#| label: all-in-one
corrected <- sw_batch_correct(
  uncorrected = uncorrected,
  markers = markers,
  covar = "condition",     # Preserve biological condition
  norm_method = "scale",   # z-score normalization; use "rank" for heavy effects
  xdim = 8,                # SOM grid dimensions
  ydim = 8,
  rlen = 10,               # SOM training iterations
  seed = 42
)

saveRDS(corrected, "corrected.rds")
```

### Key parameters

| Parameter | Default | Guidance |
|-----------|---------|----------|
| `covar` | `NULL` | Set to `"condition"` to preserve biological differences |
| `norm_method` | `"scale"` | Use `"rank"` when combining data with heavy batch effects or across platforms |
| `xdim`, `ydim` | 8 | Lower values (e.g., 6) for coarse correction; higher (e.g., 10) for fine-grained |
| `rlen` | 10 | Increase (e.g., 100) if using rank normalization or results are unconvincing |
| `seed` | 42 | For reproducibility |

## Modular batch correction workflow

The modular workflow gives you full control over each step. This is
recommended when you need to:

- Tune the normalization method independently
- Adjust SOM grid dimensions iteratively
- Use custom cluster labels (e.g., from manual gating or FlowSOM)
- Debug issues at a specific step

### Step 1: Normalize

Per-batch normalization ensures that marker distributions are comparable
before SOM clustering.

```{r}
#| label: normalize
normalized <- sw_normalize(
  df = uncorrected,
  markers = markers,
  norm_method = "scale"  # "scale" (z-score) or "rank" (rank-based)
)
```

::: {.callout-tip}
## When to Use Rank Normalization

Use `norm_method = "rank"` when combining data with heavy batch effects,
such as data from different instruments, different panels, or different
sites. For single-study data with moderate batch effects,
`norm_method = "scale"` usually suffices.
:::

### Step 2: Create SOM clusters

The SOM groups cells into phenotypically similar clusters. ComBat then
corrects batch effects **within** each cluster, preventing the loss of
real biological heterogeneity.

```{r}
#| label: create-som
labels <- sw_create_som(
  df = normalized,
  markers = markers,
  xdim = 8,
  ydim = 8,
  rlen = 10,
  seed = 42
)

# labels is an integer vector — one cluster assignment per cell
table(labels)
```

::: {.callout-note}
## SOM Grid Size Guidance

- **6×6** (36 clusters): Coarse correction for very different batches
- **8×8** (64 clusters): Good default for most spectral flow experiments
- **10×10** (100 clusters): Fine-grained correction when batch effects are subtle
:::

### Step 3: Apply ComBat correction

Pass the **original** (un-normalized) data together with the SOM labels
to ComBat. Normalization was only used for clustering — the correction
operates on the original scale.

```{r}
#| label: correct
corrected <- sw_correct_data(
  df = uncorrected,     # Original data, NOT normalized
  label = labels,       # SOM cluster labels from step 2
  markers = markers,
  covar = "condition",  # Biological covariate to preserve
  parametric = TRUE     # Parametric ComBat (default)
)

saveRDS(corrected, "corrected.rds")
```

### Using custom cluster labels

You are not restricted to SOM labels. Any integer vector of cluster
assignments (same length as `nrow(uncorrected)`) can be used — for
example, from manual gating or from a FlowSOM run with
`sw_cluster()`:

```{r}
#| label: custom-labels
# Example: use FlowSOM metaclusters as labels
# Note: sw_cluster()'s first parameter is named 'corrected' but accepts
# any expression tibble — here we intentionally pass uncorrected data
# to obtain cluster labels for use with sw_correct_data()
fsom <- sw_cluster(
  corrected = uncorrected,
  lineage_markers = markers,
  xdim = 10,
  ydim = 10,
  n_metaclusters = 30
)
custom_labels <- sw_get_cluster_assignments(fsom)

corrected_custom <- sw_correct_data(
  df = uncorrected,
  label = custom_labels,
  markers = markers,
  covar = "condition"
)
```

## Evaluate correction quality

### Quick evaluation with MAD

`sw_evaluate_correction()` provides a fast assessment using the Median
Absolute Deviation (MAD) of batch medians:

```{r}
#| label: eval-quick
eval_result <- sw_evaluate_correction(
  uncorrected = uncorrected,
  corrected = corrected,
  markers = markers
)

# Was there improvement?
eval_result$improved
# [1] TRUE

# Percentage reduction in MAD (field is named emd_reduction_pct for historical reasons)
eval_result$emd_reduction_pct
# [1] 72.3

# Per-marker MAD scores
eval_result$mad
```

### Earth Mover's Distance evaluation

For a more rigorous, cluster-aware assessment, use `sw_evaluate_emd()`.
This computes EMD within each SOM cluster before and after correction:

```{r}
#| label: eval-emd
# Both data frames need a 'label' column
uncorrected$label <- corrected$label <- labels

emd_eval <- sw_evaluate_emd(
  uncorrected = uncorrected,
  corrected = corrected,
  markers = markers,
  cell_col = "label",
  batch_col = "batch"
)

# Overall EMD reduction
emd_eval$reduction

# Violin plot comparing EMDs
emd_eval$violin

# Scatter plot of per-marker EMDs
emd_eval$scatter
```

### MAD evaluation (cyCombine-backed)

```{r}
#| label: eval-mad
mad_eval <- sw_evaluate_mad(
  uncorrected = uncorrected,
  corrected = corrected,
  markers = markers,
  cell_col = "label",
  batch_col = "batch"
)

# Overall MAD score
mad_eval$score
```

## Visualize results

### Density plots

Compare marker density distributions across batches before and after
correction:

```{r}
#| label: density-plots
sw_plot_batch_densities(
  uncorrected = uncorrected,
  corrected = corrected,
  markers = markers,
  filename = "batch_densities.pdf",
  ncol = 6
)
```

### Dimensionality reduction

UMAP or PCA projections show whether batch-driven clusters collapse
after correction:

```{r}
#| label: dimred-plots
# Subsample for speed
set.seed(42)
idx <- sample(nrow(uncorrected), min(20000, nrow(uncorrected)))

# UMAP plots
p_uncorrected <- sw_plot_batch_dimred(
  df = uncorrected[idx, ],
  markers = markers,
  type = "umap",
  name = "Uncorrected"
)

p_corrected <- sw_plot_batch_dimred(
  df = corrected[idx, ],
  markers = markers,
  type = "umap",
  name = "Corrected"
)

# Display side by side (requires patchwork or gridExtra)
library(patchwork)
p_uncorrected + p_corrected
```

You can also use PCA instead of UMAP:

```{r}
#| label: pca-plots
p_pca <- sw_plot_batch_dimred(
  df = corrected[idx, ],
  markers = markers,
  type = "pca",
  name = "Corrected (PCA)"
)
p_pca
```

## Complete example: putting it all together

Here is the full modular workflow in a single script:

```{r}
#| label: complete-example
library(SpectraWeaveR)

# ── 1. Load data ──────────────────────────────────────────────
fcs_dir <- "path/to/unmixed_fcs"
fs <- sw_load_unmixed(fcs_dir)

markers <- c("CD3", "CD4", "CD8", "CD19", "CD14", "CD56",
             "CD45RA", "CD27", "HLA-DR", "CD69")

sample_meta <- data.frame(
  sample = sampleNames(fs),
  batch = c(rep("Day1", 5), rep("Day2", 5)),
  condition = c(rep("Ctrl", 3), rep("Stim", 2),
                rep("Ctrl", 3), rep("Stim", 2))
)

ff_list <- setNames(
  lapply(sampleNames(fs), function(s) fs[[s]]),
  sampleNames(fs)
)

# ── 2. Prepare tibble ────────────────────────────────────────
uncorrected <- sw_prepare_for_correction(
  ff_list, sample_meta, markers, cofactor = 6000
)

# ── 3. Detect batch effects ──────────────────────────────────
sw_detect_batch_effect(uncorrected, markers, out_dir = "diagnostics")

# ── 4. Modular batch correction ──────────────────────────────
# Step 4a: Normalize
normalized <- sw_normalize(uncorrected, markers, norm_method = "scale")

# Step 4b: Cluster
labels <- sw_create_som(normalized, markers, xdim = 8, ydim = 8, rlen = 10)

# Step 4c: Correct
corrected <- sw_correct_data(
  uncorrected, label = labels, markers = markers, covar = "condition"
)

# ── 5. Evaluate ──────────────────────────────────────────────
eval_result <- sw_evaluate_correction(uncorrected, corrected, markers)
message("MAD reduction: ", eval_result$emd_reduction_pct, "%")

# Cluster-aware evaluation
uncorrected$label <- corrected$label <- labels
emd_eval <- sw_evaluate_emd(uncorrected, corrected, markers)
message("EMD reduction: ", emd_eval$reduction, "%")

# ── 6. Visualize ─────────────────────────────────────────────
sw_plot_batch_densities(uncorrected, corrected, markers,
                        filename = "densities.pdf")

set.seed(42)
idx <- sample(nrow(uncorrected), 20000)
p1 <- sw_plot_batch_dimred(uncorrected[idx, ], markers,
                           type = "umap", name = "Uncorrected")
p2 <- sw_plot_batch_dimred(corrected[idx, ], markers,
                           type = "umap", name = "Corrected")

# ── 7. Continue to clustering ────────────────────────────────
fsom <- sw_cluster(
  corrected = corrected,
  lineage_markers = markers,
  n_metaclusters = 20
)
sw_plot_clusters(fsom, plot_file = "FlowSOM_clusters.pdf")
```

## Tips and best practices

1. **Cofactor**: Use 6000 for spectral flow cytometry (not 5 as used in
   CyTOF). The higher cofactor matches the wider dynamic range of
   Aurora/spectral instruments (~4×10⁶).

2. **Normalization**: Start with `"scale"`. Switch to `"rank"` if you are
   combining data from very different sources (e.g., spectral flow + CyTOF).

3. **SOM grid size**: An 8×8 grid (64 clusters) is a good starting point.
   If correction seems too aggressive, increase the grid; if batch effects
   persist, decrease it.

4. **Biological covariates**: Always set `covar = "condition"` (or your
   equivalent column) to prevent ComBat from removing real biological
   differences.

5. **Evaluation**: Use both EMD and MAD metrics. A good correction shows
   EMD reduction >50% while maintaining biological separation in
   density plots.

6. **Downstream**: After batch correction, proceed to clustering with
   `sw_cluster()`. The corrected tibble can be passed directly.

## Session info

```{r}
#| label: session-info
#| eval: true
sessionInfo()
```
