---
title: "LLM-Powered Pipeline Builder"
vignette: >
  %\VignetteIndexEntry{LLM-Powered Pipeline Builder}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

SpectraWeaveR includes an optional **LLM-powered assistant** that guides
you through configuring your spectral flow cytometry analysis pipeline via
natural language conversation. The assistant is powered by
[ellmer](https://ellmer.tidyverse.org/) and can inspect your files,
validate your metadata, and generate runnable R code.

## Prerequisites

### Install ellmer

```{r}
install.packages("ellmer")
```

### Set up an API key

The assistant uses a large language model (LLM) provider. Set up an API
key for your chosen provider:

**OpenAI** (default):

```{r}
# Add to your ~/.Renviron file:
# OPENAI_API_KEY=sk-...
Sys.setenv(OPENAI_API_KEY = "sk-your-key-here")
```

**Anthropic Claude**:

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

**Local models** (no API key needed):

```{r}
# Install and start Ollama (https://ollama.ai)
# Then use provider = "ollama"
```

## Quick Start

### Interactive Session

Launch the assistant with a single call:

```{r}
library(SpectraWeaveR)
sw_assistant()
```

This starts a live conversation in your R console. The assistant will:

1. Ask about your experiment (cytometer, number of samples, batches)
2. Request paths to your FCS files and metadata
3. Inspect your files using built-in tools
4. Help you select markers and configure parameters
5. Generate a complete, runnable SpectraWeaveR pipeline

### Example Session

A typical conversation might go like this:

- **Assistant**: "What cytometer did you use and how many samples do you have?"
- **You**: "Aurora 5L, 30 samples across 3 batches"
- **Assistant**: "Where are your FCS files located?"
- **You**: "/data/experiment1/fcs/"
- *The assistant inspects the directory and lists found FCS files*
- **Assistant**: "I found 30 FCS files. Do you have a sample metadata file?"
- **You**: "/data/experiment1/metadata.csv"
- *The assistant reads the CSV, shows columns, and validates against FCS files*
- **Assistant**: "Which column contains the batch labels?"
- **You**: "batch_number"
- *The assistant continues until all parameters are configured, then generates code*

### Programmatic Use

You can also get the `Chat` object for programmatic interaction:

```{r}
chat <- sw_assistant(interactive = FALSE)
response <- chat$chat("I have 20 Aurora FCS files in /data/fcs/")
```

## Specialized Assistants

SpectraWeaveR provides focused assistants for specific tasks:

### Batch Correction

```{r}
sw_assistant_batch_correction()
```

Helps configure batch correction: metadata validation, marker selection,
normalization method, and evaluation code.

### Gating

```{r}
sw_assistant_gating()
```

Helps create gating templates and apply automated gating with openCyto.

### Spectral Unmixing

```{r}
sw_assistant_unmixing()
```

Helps configure the AutoSpectral unmixing workflow from raw spectral data.

## One-Shot Pipeline Generation

For scripts that need LLM-generated code without interactive conversation:

```{r}
code <- sw_quick_pipeline(
  description = "I have 40-color Aurora 5L data, 30 samples across 3 batches,
    FCS files in /data/fcs/, metadata in /data/meta.csv with columns:
    filename, patient_id, batch_number, treatment_group.
    Cluster on lineage markers CD3, CD4, CD8, CD19, CD56.",
  provider = "openai"
)
cat(code)
```

## Code Generation from Config

You can also generate code from a configuration list without LLM:

```{r}
config <- list(
  fcs_dir = "/data/fcs",
  sample_meta_path = "/data/metadata.csv",
  markers = c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14",
              "CD16", "HLA-DR", "CD45RA", "CD45RO"),
  lineage_markers = c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14"),
  batch_col = "batch_number",
  sample_col = "patient_id",
  file_col = "filename",
  condition_col = "treatment_group",
  cofactor = 6000,
  n_metaclusters = 20,
  output_dir = "results",
  seed = 42
)

# Print to console
sw_generate_pipeline_code(config)

# Save to file
sw_generate_pipeline_code(config, output = "file", path = "my_pipeline.R")

# Composable pipeline style
sw_generate_pipeline_code(config, style = "composable")
```

## Using Different Providers

### OpenAI (default)

```{r}
sw_assistant(provider = "openai")
sw_assistant(provider = "openai", model = "gpt-4o")
```

### Anthropic Claude

```{r}
sw_assistant(provider = "anthropic")
sw_assistant(provider = "anthropic", model = "claude-sonnet-4-20250514")
```

### Google Gemini

```{r}
sw_assistant(provider = "google")
```

### Ollama (Local)

For fully private operation with no data sent externally:

```{r}
# First install and start Ollama, then pull a model:
# ollama pull llama3
sw_assistant(provider = "ollama", model = "llama3")
```

## Structured Configuration Extraction

Advanced users can use ellmer's structured data extraction to get a
validated configuration object from a conversation:
```{r}
chat <- sw_assistant(interactive = FALSE)
# ... have a conversation ...

config_type <- sw_pipeline_config_type()
config <- chat$chat_structured(
  "Extract the pipeline configuration from our conversation.",
  type = config_type
)

# Generate code from the extracted config
sw_generate_pipeline_code(config)
```

## Privacy and Security

The assistant sends the following to the LLM provider:

- File and directory paths
- Directory listings (file names only)
- FCS file headers (channel names, not expression data)
- CSV/XLSX column names and preview rows
- Your conversation messages

**No cell expression data** (fluorescence intensities) is transmitted.

For fully local operation, use Ollama:

```{r}
sw_assistant(provider = "ollama", model = "llama3")
```

To suppress the privacy notice:

```{r}
sw_assistant(privacy_notice = FALSE)
```