---
title: "LLM-Powered Pipeline Builder"
vignette: >
  %\VignetteIndexEntry{LLM-Powered Pipeline Builder}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

SpectraWeaveR includes an optional **LLM-powered assistant** that guides
you through configuring your spectral flow cytometry analysis pipeline via
natural language conversation. The assistant is powered by
[ellmer](https://ellmer.tidyverse.org/) and can inspect your files,
validate your metadata, and generate runnable R code.

## Prerequisites

### Install ellmer

```{r}
install.packages("ellmer")
```

### Set up an API key

The assistant uses a large language model (LLM) provider. Set up an API
key for your chosen provider:

**OpenAI** (default):

```{r}
# Add to your ~/.Renviron file:
# OPENAI_API_KEY=sk-...
Sys.setenv(OPENAI_API_KEY = "sk-your-key-here")
```

**Anthropic Claude**:

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

**Local models** (no API key needed):

```{r}
# Install and start Ollama (https://ollama.ai)
# Then use provider = "ollama"
```

## Quick Start

### Interactive Session

Launch the assistant with a single call:

```{r}
library(SpectraWeaveR)
sw_assistant()
```

This starts a live conversation in your R console. The assistant will:

1. Ask about your experiment (cytometer, number of samples, batches)
2. Request paths to your FCS files and metadata
3. Inspect your files using built-in tools
4. Help you select markers and configure parameters
5. Generate a complete, runnable SpectraWeaveR pipeline

### Example Session

A typical conversation might go like this:

- **Assistant**: "What cytometer did you use and how many samples do you have?"
- **You**: "Aurora 5L, 30 samples across 3 batches"
- **Assistant**: "Where are your FCS files located?"
- **You**: "/data/experiment1/fcs/"
- *The assistant inspects the directory and lists found FCS files*
- **Assistant**: "I found 30 FCS files. Do you have a sample metadata file?"
- **You**: "/data/experiment1/metadata.csv"
- *The assistant reads the CSV, shows columns, and validates against FCS files*
- **Assistant**: "Which column contains the batch labels?"
- **You**: "batch_number"
- *The assistant continues until all parameters are configured, then generates code*

### Programmatic Use

You can also get the `Chat` object for programmatic interaction:

```{r}
chat <- sw_assistant(interactive = FALSE)
response <- chat$chat("I have 20 Aurora FCS files in /data/fcs/")
```

## Specialized Assistants

SpectraWeaveR provides focused assistants for specific tasks:

### Batch Correction

```{r}
sw_assistant_batch_correction()
```

Helps configure batch correction: metadata validation, marker selection,
normalization method, and evaluation code.

### Gating

```{r}
sw_assistant_gating()
```

Helps create gating templates and apply automated gating with openCyto.

### Spectral Unmixing

```{r}
sw_assistant_unmixing()
```

Helps configure the AutoSpectral unmixing workflow from raw spectral data.

## One-Shot Pipeline Generation

For scripts that need LLM-generated code without interactive conversation:

```{r}
code <- sw_pipeline_quick(
  description = "I have 40-color Aurora 5L data, 30 samples across 3 batches,
    FCS files in /data/fcs/, metadata in /data/meta.csv with columns:
    filename, patient_id, batch_number, treatment_group.
    Cluster on lineage markers CD3, CD4, CD8, CD19, CD56.",
  provider = "openai"
)
cat(code)
```

## Code Generation from Config

You can also generate code from a configuration list without LLM:

```{r}
config <- list(
  fcs_dir = "/data/fcs",
  sample_meta_path = "/data/metadata.csv",
  markers = c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14",
              "CD16", "HLA-DR", "CD45RA", "CD45RO"),
  lineage_markers = c("CD3", "CD4", "CD8", "CD19", "CD56", "CD14"),
  batch_col = "batch_number",
  sample_col = "patient_id",
  file_col = "filename",
  condition_col = "treatment_group",
  cofactor = 6000,
  n_metaclusters = 20,
  output_dir = "results",
  seed = 42
)

# Print to console
sw_pipeline_generate_code(config)

# Save to file
sw_pipeline_generate_code(config, output = "file", path = "my_pipeline.R")

# Composable pipeline style
sw_pipeline_generate_code(config, style = "composable")
```

## Using Different Providers

### OpenAI (default)

```{r}
sw_assistant(provider = "openai")
sw_assistant(provider = "openai", model = "gpt-4o")
```

### Anthropic Claude

```{r}
sw_assistant(provider = "anthropic")
sw_assistant(provider = "anthropic", model = "claude-sonnet-4-20250514")
```

### Google Gemini

```{r}
sw_assistant(provider = "google")
```

### Ollama (Local)

For fully private operation with no data sent externally:

```{r}
# First install and start Ollama, then pull a model:
# ollama pull llama3
sw_assistant(provider = "ollama", model = "llama3")
```

## Structured Configuration Extraction

Advanced users can use ellmer's structured data extraction to get a
validated configuration object from a conversation:
```{r}
chat <- sw_assistant(interactive = FALSE)
# ... have a conversation ...

config_type <- sw_pipeline_config_type()
config <- chat$chat_structured(
  "Extract the pipeline configuration from our conversation.",
  type = config_type
)

# Generate code from the extracted config
sw_pipeline_generate_code(config)
```

## R Environment Introspection (btw)

When the [btw](https://posit-dev.github.io/btw/) package is installed, the
assistant automatically gains the ability to inspect your R environment. This
means it can:

- **Describe data frames** in your session (column names, types, summary
  statistics)
- **Read package documentation** (help pages, vignettes)
- **Inspect session info** (loaded packages, R version)
- **Explore local files** in your working directory

Install btw:

```{r}
install.packages("btw")
```

The btw tools are registered automatically when available. To disable them:

```{r}
sw_assistant(include_btw = FALSE)
```

## MCP Server for External AI Tools

SpectraWeaveR can expose its file-inspection tools via the
[Model Context Protocol](https://modelcontextprotocol.io) (MCP), allowing
external AI applications — Claude Code, VS Code Copilot Chat, Cursor — to
call them directly.

### Setup

Install the required packages:

```{r}
install.packages(c("mcptools", "btw"))
```

### Claude Code

```bash
claude mcp add -s user spectraweaver -- \
  Rscript -e "SpectraWeaveR::sw_assistant_mcp()"
```

### VS Code (`.vscode/mcp.json`)

```json
{
  "mcpServers": {
    "spectraweaver": {
      "command": "Rscript",
      "args": ["-e", "SpectraWeaveR::sw_assistant_mcp()"]
    }
  }
}
```

Once configured, your AI coding assistant can inspect FCS files, validate
metadata, detect channels, and check batch balance without leaving your
editor.

## Privacy and Security

### What is sent to the LLM provider

- File and directory paths
- Directory listings (file names only)
- FCS file headers (channel names, not expression data)
- CSV/XLSX column names and preview rows
- Your conversation messages

**No cell expression data** (fluorescence intensities) is transmitted.

### Sensitivity levels

SpectraWeaveR provides two privacy modes via the `sensitivity` parameter:

| Aspect | `"standard"` (default) | `"strict"` |
|--------|----------------------|------------|
| CSV/XLSX preview | Auto-redacts PII columns | Column names only (no row values) |
| btw tool groups | env, docs, session, files | docs, session only |
| Batch cross-tabs | Shown | Shown with advisory note |
| FCS headers | Channel names only (safe) | Channel names only (safe) |

PII detection matches common patterns: `patient_id`, `name`, `dob`, `mrn`,
`ssn`, `diagnosis`, `address`, `email`, etc.

### For clinical / sensitive data

```{r}
# Maximum privacy: local model + strict mode
sw_assistant(provider = "ollama", model = "llama3",
             sensitivity = "strict")

# Or just strict mode with a cloud provider
sw_assistant(sensitivity = "strict")
```

### Other privacy controls

```{r}
# Disable btw environment tools entirely
sw_assistant(include_btw = FALSE)

# Suppress the privacy notice
sw_assistant(privacy_notice = FALSE)
```