---
title: "Spectral Unmixing with SpectraWeaveR and AutoSpectral"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spectral Unmixing with SpectraWeaveR and AutoSpectral}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

SpectraWeaveR provides a streamlined interface to the
[AutoSpectral](https://github.com/DrCytometer/AutoSpectral) package for
spectral unmixing of high-parameter flow cytometry data. AutoSpectral's
full workflow normally requires 10+ individual function calls with complex
intermediate data structures. SpectraWeaveR wraps these into **six functions** — five step-by-step
functions plus a one-call orchestrator — making spectral
unmixing accessible while preserving full expert control.

### The six unmixing functions

| Function | Purpose | AutoSpectral functions wrapped |
|---|---|---|
| `sw_autospectral_setup()` | Initialize parameters, create/validate control file | `get.autospectral.param()`, `create.control.file()`, `check.control.file()` |
| `sw_prepare_controls()` | Gate, clean, extract fluorophore spectra | `define.flow.control()`, `clean.controls()`, `get.fluorophore.spectra()` |
| `sw_extract_af_spectra()` | Per-cell autofluorescence extraction | `get.af.spectra()` |
| `sw_extract_spectral_variants()` | Map fluorophore emission variability | `get.spectral.variants()` |
| `sw_unmix()` | Unmix fully-stained samples | `unmix.fcs()`, `unmix.folder()` |
| `sw_unmix_pipeline()` | One-call end-to-end orchestrator | All of the above |

## Prerequisites

### Install AutoSpectral

AutoSpectral is a GitHub-only package (not on CRAN or Bioconductor):

```{r install-autospectral}
# Required
remotes::install_github("DrCytometer/AutoSpectral")

# Strongly recommended for speed (C++ acceleration)
remotes::install_github("DrCytometer/AutoSpectralRcpp")
```

### Required data

To follow along, you will need:

1. **Single-stain control FCS files** — one file per fluorophore, plus an
   unstained control and any compensation beads. These go in a single
   directory (the "control directory").
2. **Unstained sample FCS file(s)** — for autofluorescence extraction.
   For multi-tissue experiments, one unstained file per tissue type.
3. **Fully-stained sample FCS files** — your experimental data to unmix.

## Step-by-step workflow

### Step 1: Setup — `sw_autospectral_setup()`

Initialize AutoSpectral parameters for your cytometer and create (or
validate) the control definition CSV file.

```{r step1-setup}
library(SpectraWeaveR)

# Point to your single-stain controls directory
control_dir <- "path/to/controls/"

# Initialize setup for a Cytek Aurora
setup <- sw_autospectral_setup(
  control_dir = control_dir,
  cytometer    = "aurora",
  output_dir   = "unmix_output"
)
```

When `control_file = NULL` (the default), a draft CSV is auto-generated
from the FCS filenames. **You must review and edit this file** before
proceeding — it maps each FCS file to its fluorophore name and tells
AutoSpectral which files are unstained controls, beads, etc.

If you already have a validated control file:

```{r step1-existing-csv}
setup <- sw_autospectral_setup(
  control_dir  = control_dir,
  cytometer    = "aurora",
  control_file = "path/to/my_control_file.csv",
  output_dir   = "unmix_output"
)
```

The returned `setup` object (class `sw_setup`) contains the AutoSpectral
parameter list (`asp`), the validated control file path, and the control
directory. Pass it to all subsequent functions.

**Supported cytometers:** `"aurora"` (default), `"auroraNL"`, `"id7000"`,
`"s8"`, `"a8"`, `"a5se"`, `"opteon"`, `"mosaic"`, `"xenith"`.

### Step 2: Prepare controls — `sw_prepare_controls()`

Load the control FCS files, apply automated gating, clean the controls,
and extract the fluorophore spectral signatures.

```{r step2-controls}
controls <- sw_prepare_controls(
  setup         = setup,
  gating_system = "landmarks",  # or "density"
  clean         = TRUE,         # remove AF intrusions, universal negatives
  af_remove     = TRUE,
  parallel      = FALSE
)

# What's in the result?
names(controls)
#> [1] "flow_control" "spectra" "setup"

# The spectra matrix: fluorophores in rows, detectors in columns
dim(controls$spectra)
#> [1]  40  64   # e.g., 40 fluorophores across 64 detectors
```

**Expert escape hatch:** To customize gating, pre-define your gates with
`AutoSpectral::define.gate.landmarks()` and pass them via `gate_list`:

```{r step2-custom-gates}
# Define custom gates (expert users)
my_gates <- AutoSpectral::define.gate.landmarks(
  control_dir, setup$control_file, setup$asp
)

controls <- sw_prepare_controls(
  setup     = setup,
  gate_list = my_gates  # use pre-defined gates
)
```

### Step 3: Extract autofluorescence spectra — `sw_extract_af_spectra()`

Capture per-cell autofluorescence variation from unstained samples using
Self-Organizing Map (SOM) clustering. This is what gives AutoSpectral its
edge over conventional unmixing — it models cell-to-cell AF variation
instead of using a single AF spectrum.

```{r step3-af}
# Single unstained file
af_spectra <- sw_extract_af_spectra(
  unstained_fcs = "path/to/unstained.fcs",
  setup         = setup,
  spectra       = controls$spectra,
  refine        = TRUE,   # second pass for complex AF
  som_dim       = 10      # up to 100 AF clusters
)

dim(af_spectra)
#> [1] 100  64   # 100 AF clusters across 64 detectors
```

**Multi-tissue experiments:** For different tissue types with different
autofluorescence profiles, provide a named list of unstained files:

```{r step3-multitissue}
af_spectra <- sw_extract_af_spectra(
  unstained_fcs = list(
    spleen = "path/to/spleen_unstained.fcs",
    lung   = "path/to/lung_unstained.fcs",
    liver  = "path/to/liver_unstained.fcs"
  ),
  setup   = setup,
  spectra = controls$spectra
)

# Returns a named list — one AF matrix per tissue
names(af_spectra)
#> [1] "spleen" "lung" "liver"
```

### Step 4: Extract spectral variants — `sw_extract_spectral_variants()`

Map fluorophore emission variability from single-stain controls. This
captures spectral variation (e.g., tandem dye degradation) that enables
per-cell fluorophore optimization during unmixing.

This step is **optional but recommended** — it can significantly reduce
spillover spread.

```{r step4-variants}
# For single-tissue AF, pass the af_spectra matrix directly
variants <- sw_extract_spectral_variants(
  setup      = setup,
  spectra    = controls$spectra,
  af_spectra = af_spectra,  # or af_spectra$spleen for multi-tissue
  refine     = TRUE,
  som_dim    = 10,
  n_cells    = 2000  # brightest cells per fluorophore
)

# Returns an sw_variants object with:
# - variants$thresholds   : positivity thresholds
# - variants$variants     : named list of variant matrices
# - variants$delta.list   : pre-computed spectral deltas
# - variants$delta.norms  : pre-computed delta norms
```

### Step 5: Unmix samples — `sw_unmix()`

Perform spectral unmixing on your fully-stained experimental FCS files.

```{r step5-unmix}
result <- sw_unmix(
  input            = "path/to/samples/",  # directory of FCS files
  spectra          = controls$spectra,
  setup            = setup,
  flow_control     = controls$flow_control,
  method           = "AutoSpectral",
  af_spectra       = af_spectra,
  spectra_variants = variants,
  speed            = "fast",     # "fast", "medium", or "slow"
  parallel         = TRUE,
  chunk_size       = 2e6         # events per chunk (memory management)
)

# Where are the unmixed files?
result$output_files
#> [1] "unmix_output/unmixed_fcs/sample1_AutoSpectral.fcs"
#> [2] "unmix_output/unmixed_fcs/sample2_AutoSpectral.fcs"
```

The `input` parameter is flexible:

```{r step5-input-types}
# A directory — unmixes all FCS files in it
result <- sw_unmix(input = "path/to/samples/", ...)

# A single file
result <- sw_unmix(input = "path/to/sample1.fcs", ...)

# Multiple specific files
result <- sw_unmix(input = c("sample1.fcs", "sample3.fcs"), ...)
```

**Unmixing methods:**

| Method | Description | Requires AF? | Per-cell optimization? |
|---|---|---|---|
| `"AutoSpectral"` | Full per-cell unmixing (default) | Yes | Yes |
| `"OLS"` | Ordinary least squares | No | No |
| `"WLS"` | Weighted least squares | No | No |
| `"Poisson"` | Poisson regression | No | No |
| `"Automatic"` | AutoSpectral auto-selects | Yes | Yes |

## One-call pipeline: `sw_unmix_pipeline()`

For a quick end-to-end unmixing without managing intermediate objects, use
the orchestrator:

```{r pipeline-one-call}
result <- sw_unmix_pipeline(
  control_dir   = "path/to/controls/",
  sample_input  = "path/to/samples/",
  unstained_fcs = "path/to/unstained.fcs",
  cytometer     = "aurora",
  control_file  = "path/to/control_file.csv",
  method        = "AutoSpectral",
  speed         = "fast",
  refine        = TRUE,
  output_dir    = "unmix_output",
  parallel      = FALSE
)

# Access all intermediate results
names(result)
#> [1] "setup" "flow_control" "spectra" "af_spectra"
#> [5] "spectra_variants" "unmixed"

# Unmixed files
result$unmixed$output_files
```

The pipeline:

1. Calls `sw_autospectral_setup()` (Step 1)
2. Calls `sw_prepare_controls()` (Step 2)
3. Calls `sw_extract_af_spectra()` (Step 3) — skipped for OLS/WLS
4. Calls `sw_extract_spectral_variants()` (Step 4) — skipped for OLS/WLS
5. Calls `sw_unmix()` (Step 5)

### Simpler unmixing with OLS/WLS

If you don't need per-cell optimization, you can use a simpler method.
Steps 3–4 are skipped automatically:

```{r pipeline-ols}
result <- sw_unmix_pipeline(
  control_dir  = "path/to/controls/",
  sample_input = "path/to/samples/",
  cytometer    = "aurora",
  control_file = "path/to/control_file.csv",
  method       = "OLS",
  output_dir   = "unmix_output"
)
# No unstained_fcs needed — AF extraction is skipped
```

## Connecting to the downstream pipeline

After unmixing, the output FCS files can flow directly into the rest of
the SpectraWeaveR pipeline (gating → QC → batch correction → clustering):

```{r downstream}
# Load the unmixed files
fs <- sw_load_unmixed(result$unmixed$output_dir)

# Remove margin events (must be done BEFORE transformation)
ff_list <- lapply(seq_along(fs), function(i) {
  sw_remove_margins(fs[[i]])
})
names(ff_list) <- flowCore::sampleNames(fs)

# Continue with the rest of the pipeline...
# sw_signal_qc_batch(ff_list, ...)
# sw_batch_correct(...)
# sw_cluster(...)
```

Or use `run_pipeline()` with `unmix_from_raw = TRUE` to run everything
from raw spectral data through to clustering in a single call:

```{r full-pipeline}
results <- run_pipeline(
  fcs_dir         = "path/to/samples/",     # raw spectral FCS
  sample_meta     = sample_meta,
  markers         = markers,
  lineage_markers = lineage_markers,
  unmix_from_raw  = TRUE,                   # enable unmixing step
  control_dir     = "path/to/controls/",
  unstained_fcs   = "path/to/unstained.fcs",
  cytometer       = "aurora",
  unmix_method    = "AutoSpectral"
)
```

## Multi-tissue unmixing

For experiments with multiple tissue types (e.g., spleen, lung, liver),
each tissue has different autofluorescence. The step-by-step workflow
handles this:

```{r multi-tissue}
# Steps 1-2: Same for all tissues (controls are the same)
setup    <- sw_autospectral_setup(control_dir, cytometer = "aurora")
controls <- sw_prepare_controls(setup)

# Step 3: Extract tissue-specific AF
af_spectra <- sw_extract_af_spectra(
  unstained_fcs = list(
    spleen = "spleen_unstained.fcs",
    lung   = "lung_unstained.fcs"
  ),
  setup   = setup,
  spectra = controls$spectra
)

# Step 4: Extract variants using one tissue's AF (controls match one tissue)
variants <- sw_extract_spectral_variants(
  setup      = setup,
  spectra    = controls$spectra,
  af_spectra = af_spectra$spleen  # use AF matching control tissue
)

# Step 5: Unmix each tissue separately with matched AF
spleen_result <- sw_unmix(
  input        = "path/to/spleen_samples/",
  spectra      = controls$spectra,
  setup        = setup,
  flow_control = controls$flow_control,
  af_spectra   = af_spectra$spleen,
  spectra_variants = variants
)

lung_result <- sw_unmix(
  input        = "path/to/lung_samples/",
  spectra      = controls$spectra,
  setup        = setup,
  flow_control = controls$flow_control,
  af_spectra   = af_spectra$lung,
  spectra_variants = variants
)
```

## Tips and best practices

### Performance

- **Install AutoSpectralRcpp** for a significant speed improvement in
  AF extraction and unmixing.
- Use `parallel = TRUE` and set `threads` to the number of available
  cores for large experiments.
- The `speed` parameter in `sw_unmix()` controls the precision-speed
  trade-off: `"fast"` tests 1 variant per cell, `"medium"` tests 3, and
  `"slow"` tests 10.
- Adjust `chunk_size` (default `2e6`) for memory management on
  constrained systems.

### Quality control

- Always review the auto-generated control file before proceeding from
  Step 1 to Step 2.
- Check the QC figures generated in the output directory — they show
  gating, spectral signatures, and similarity heatmaps.
- Use `refine = TRUE` (the default) for tissue samples with complex
  autofluorescence — the second-pass refinement improves AF modeling.

### Reproducibility

- All intermediate objects are preserved in the return values for full
  provenance tracking.
- The `sw_unmix_pipeline()` orchestrator stores the complete chain of
  results for reproducibility.
- Save intermediate results (e.g., `spectra`, `af_spectra`, `variants`)
  as RDS files for reuse across sessions:

```{r save-intermediates}
saveRDS(controls$spectra, "spectra.rds")
saveRDS(af_spectra, "af_spectra.rds")
saveRDS(variants, "spectral_variants.rds")

# Reload in a new session
spectra  <- readRDS("spectra.rds")
af_specs <- readRDS("af_spectra.rds")
vars     <- readRDS("spectral_variants.rds")
```

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
