---
title: "Spectral Unmixing with SpectraWeaveR and AutoSpectral"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spectral Unmixing with SpectraWeaveR and AutoSpectral}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

SpectraWeaveR provides a streamlined interface to the
[AutoSpectral](https://github.com/DrCytometer/AutoSpectral) package for
spectral unmixing of high-parameter flow cytometry data. AutoSpectral's
full workflow normally requires 10+ individual function calls with complex
intermediate data structures. SpectraWeaveR wraps these into **six functions** — five step-by-step
functions plus a one-call orchestrator — making spectral
unmixing accessible while preserving full expert control.

This vignette walks through the complete unmixing workflow using the
example dataset from the
[original AutoSpectral documentation](https://drcytometer.github.io/AutoSpectral/articles/01_Full_AutoSpectral_Workflow.html).

### The six unmixing functions

| Function | Purpose | AutoSpectral functions wrapped |
|---|---|---|
| `sw_autospectral_setup()` | Initialize parameters, create/validate control file | `get.autospectral.param()`, `create.control.file()`, `check.control.file()` |
| `sw_prepare_controls()` | Gate, clean, extract fluorophore spectra | `define.flow.control()`, `clean.controls()`, `get.fluorophore.spectra()` |
| `sw_extract_af_spectra()` | Per-cell autofluorescence extraction | `get.af.spectra()` |
| `sw_extract_spectral_variants()` | Map fluorophore emission variability | `get.spectral.variants()` |
| `sw_unmix()` | Unmix fully-stained samples | `unmix.fcs()`, `unmix.folder()` |
| `sw_unmix_pipeline()` | One-call end-to-end orchestrator | All of the above |

## Prerequisites

### Install AutoSpectral

AutoSpectral is a GitHub-only package (not on CRAN or Bioconductor):

```{r install-autospectral}
# Required
remotes::install_github("DrCytometer/AutoSpectral")

# Strongly recommended for speed (C++ acceleration)
remotes::install_github("DrCytometer/AutoSpectralRcpp")
```

### Example data

This vignette uses the **9-colour Aurora panel** from the original
AutoSpectral workflow. The data are from a mouse experiment with samples
from three tissue types — **spleen, lung, and liver** — run on a
**5-laser Cytek Aurora**. The different tissues allow us to demonstrate
how to handle diverse autofluorescence profiles.

Download the dataset from
[Mendeley Data](https://data.mendeley.com/datasets/ch5dnspd79/1)
(Manurung, 2024). After extracting, you should see a directory structure
like this:

```
AutoSpectral_data/
├── Raw/
│   └── Set1/
│       ├── Reference Group/    # single-stain controls (beads + cells)
│       ├── Unstained/          # one unstained FCS per tissue
│       │   ├── D1 Spleen_Set1.fcs
│       │   ├── D2 Lung_Set1.fcs
│       │   └── D3 Liver_Set1.fcs
│       └── Stained/            # fully-stained samples
│           ├── D4 Spleen_Set1.fcs
│           ├── D5 Lung_Set1.fcs
│           └── D6 Liver_Set1.fcs
└── fcs_control_file.csv        # pre-filled control definition file
```

Set the base path to wherever you extracted the data:

```{r data-paths}
base_dir <- "~/AutoSpectral_data"
```

## Step-by-step workflow

### Step 1: Setup — `sw_autospectral_setup()`

Initialize AutoSpectral parameters for your cytometer and create (or
validate) the control definition CSV file.

```{r step1-setup}
library(SpectraWeaveR)

# Point to the single-stain controls directory
control_dir <- file.path(base_dir, "Raw", "Set1", "Reference Group")

# Initialize setup for the Cytek Aurora
setup <- sw_autospectral_setup(
  control_dir = control_dir,
  cytometer    = "aurora",
  output_dir   = "unmix_output"
)
```

When `control_file = NULL` (the default), a draft CSV is auto-generated
from the FCS filenames. **You must review and edit this file** before
proceeding — it maps each FCS file to its fluorophore name and tells
AutoSpectral which files are unstained controls, beads, etc.

The example dataset includes a pre-filled control file. To use it:

```{r step1-existing-csv}
control_file <- file.path(base_dir, "fcs_control_file.csv")

setup <- sw_autospectral_setup(
  control_dir  = control_dir,
  cytometer    = "aurora",
  control_file = control_file,
  output_dir   = "unmix_output"
)
```

The returned `setup` object (class `sw_setup`) contains the AutoSpectral
parameter list (`asp`), the validated control file path, and the control
directory. Pass it to all subsequent functions.

**Supported cytometers:** `"aurora"` (default), `"auroraNL"`, `"id7000"`,
`"s8"`, `"a8"`, `"a5se"`, `"opteon"`, `"mosaic"`, `"xenith"`.

### Step 2: Prepare controls — `sw_prepare_controls()`

Load the control FCS files, apply automated gating, clean the controls,
and extract the fluorophore spectral signatures.

```{r step2-controls}
controls <- sw_prepare_controls(
  setup         = setup,
  gating_system = "landmarks",  # or "density"
  clean         = TRUE,         # remove AF intrusions, universal negatives
  af_remove     = TRUE,
  parallel      = FALSE
)

# What's in the result?
names(controls)
#> [1] "flow_control" "spectra" "setup"

# The spectra matrix: fluorophores in rows, detectors in columns
dim(controls$spectra)
#> [1]  10  64   # 9 fluorophores + AF across 64 detectors
```

**Expert escape hatch:** To customize gating, pre-define your gates with
`AutoSpectral::define.gate.landmarks()` and pass them via `gate_list`:

```{r step2-custom-gates}
# Define custom gates (expert users)
my_gates <- AutoSpectral::define.gate.landmarks(
  control_dir, setup$control_file, setup$asp
)

controls <- sw_prepare_controls(
  setup     = setup,
  gate_list = my_gates  # use pre-defined gates
)
```

### Step 3: Extract autofluorescence spectra — `sw_extract_af_spectra()`

Capture per-cell autofluorescence variation from unstained samples using
Self-Organizing Map (SOM) clustering. This is what gives AutoSpectral its
edge over conventional unmixing — it models cell-to-cell AF variation
instead of using a single AF spectrum.

Since the example dataset has three tissue types, each with distinct
autofluorescence profiles, we extract AF spectra from each:

```{r step3-af-multi}
unstained_dir <- file.path(base_dir, "Raw", "Set1", "Unstained")

af_spectra <- sw_extract_af_spectra(
  unstained_fcs = list(
    spleen = file.path(unstained_dir, "D1 Spleen_Set1.fcs"),
    lung   = file.path(unstained_dir, "D2 Lung_Set1.fcs"),
    liver  = file.path(unstained_dir, "D3 Liver_Set1.fcs")
  ),
  setup   = setup,
  spectra = controls$spectra,
  refine  = TRUE   # second pass for complex tissue AF
)

# Returns a named list — one AF matrix per tissue
names(af_spectra)
#> [1] "spleen" "lung" "liver"
```

For single-tissue experiments (e.g., PBMCs), pass a single file path
instead of a list:

```{r step3-af-single}
af_spectra_single <- sw_extract_af_spectra(
  unstained_fcs = file.path(unstained_dir, "D1 Spleen_Set1.fcs"),
  setup         = setup,
  spectra       = controls$spectra,
  refine        = TRUE,
  som_dim       = 10   # up to 100 AF clusters
)

dim(af_spectra_single)
#> [1] 100  64   # 100 AF clusters across 64 detectors
```

### Step 4: Extract spectral variants — `sw_extract_spectral_variants()`

Map fluorophore emission variability from single-stain controls. This
captures spectral variation (e.g., tandem dye degradation) that enables
per-cell fluorophore optimization during unmixing.

This step is **optional but recommended** — it can significantly reduce
spillover spread.

We use the spleen AF spectra here because the control samples are from
spleen tissue. Always match the AF to whatever tissue your controls
were prepared from.

```{r step4-variants}
variants <- sw_extract_spectral_variants(
  setup      = setup,
  spectra    = controls$spectra,
  af_spectra = af_spectra$spleen,  # match AF to control tissue
  refine     = TRUE,
  som_dim    = 10,
  n_cells    = 2000  # brightest cells per fluorophore
)

# Returns an sw_variants object with:
# - variants$thresholds   : positivity thresholds
# - variants$variants     : named list of variant matrices
# - variants$delta.list   : pre-computed spectral deltas
# - variants$delta.norms  : pre-computed delta norms
```

### Step 5: Unmix samples — `sw_unmix()`

Perform spectral unmixing on your fully-stained experimental FCS files.
For multi-tissue experiments, unmix each tissue separately with its
matched AF spectra.

```{r step5-unmix}
stained_dir <- file.path(base_dir, "Raw", "Set1", "Stained")

# Unmix spleen samples with spleen AF
spleen_result <- sw_unmix(
  input            = file.path(stained_dir, "D4 Spleen_Set1.fcs"),
  spectra          = controls$spectra,
  setup            = setup,
  flow_control     = controls$flow_control,
  method           = "AutoSpectral",
  af_spectra       = af_spectra$spleen,
  spectra_variants = variants,
  speed            = "fast",
  parallel         = FALSE
)

# Unmix lung samples with lung AF
lung_result <- sw_unmix(
  input            = file.path(stained_dir, "D5 Lung_Set1.fcs"),
  spectra          = controls$spectra,
  setup            = setup,
  flow_control     = controls$flow_control,
  method           = "AutoSpectral",
  af_spectra       = af_spectra$lung,
  spectra_variants = variants,
  speed            = "fast",
  parallel         = FALSE
)

# Unmix liver samples with liver AF
liver_result <- sw_unmix(
  input            = file.path(stained_dir, "D6 Liver_Set1.fcs"),
  spectra          = controls$spectra,
  setup            = setup,
  flow_control     = controls$flow_control,
  method           = "AutoSpectral",
  af_spectra       = af_spectra$liver,
  spectra_variants = variants,
  speed            = "fast",
  parallel         = FALSE
)

# Where are the unmixed files?
spleen_result$output_files
#> [1] "unmix_output/unmixed_fcs/D4 Spleen_Set1_AutoSpectral.fcs"
```

The `input` parameter is flexible:

```{r step5-input-types}
# A directory — unmixes all FCS files in it
result <- sw_unmix(input = stained_dir, ...)

# A single file
result <- sw_unmix(input = file.path(stained_dir, "D4 Spleen_Set1.fcs"), ...)

# Multiple specific files
result <- sw_unmix(
  input = c(
    file.path(stained_dir, "D4 Spleen_Set1.fcs"),
    file.path(stained_dir, "D5 Lung_Set1.fcs")
  ),
  ...
)
```

**Unmixing methods:**

| Method | Description | Requires AF? | Per-cell optimization? |
|---|---|---|---|
| `"AutoSpectral"` | Full per-cell unmixing (default) | Yes | Yes |
| `"OLS"` | Ordinary least squares | No | No |
| `"WLS"` | Weighted least squares | No | No |
| `"Poisson"` | Poisson regression | No | No |
| `"Automatic"` | AutoSpectral auto-selects | Yes | Yes |

## One-call pipeline: `sw_unmix_pipeline()`

For a quick end-to-end unmixing without managing intermediate objects, use
the orchestrator. Here we unmix the spleen sample in a single call:

```{r pipeline-one-call}
result <- sw_unmix_pipeline(
  control_dir   = file.path(base_dir, "Raw", "Set1", "Reference Group"),
  sample_input  = file.path(base_dir, "Raw", "Set1", "Stained",
                            "D4 Spleen_Set1.fcs"),
  unstained_fcs = file.path(base_dir, "Raw", "Set1", "Unstained",
                            "D1 Spleen_Set1.fcs"),
  cytometer     = "aurora",
  control_file  = file.path(base_dir, "fcs_control_file.csv"),
  method        = "AutoSpectral",
  speed         = "fast",
  refine        = TRUE,
  output_dir    = "unmix_output",
  parallel      = FALSE
)

# Access all intermediate results
names(result)
#> [1] "setup" "flow_control" "spectra" "af_spectra"
#> [5] "spectra_variants" "unmixed"

# Unmixed files
result$unmixed$output_files
```

The pipeline:

1. Calls `sw_autospectral_setup()` (Step 1)
2. Calls `sw_prepare_controls()` (Step 2)
3. Calls `sw_extract_af_spectra()` (Step 3) — skipped for OLS/WLS
4. Calls `sw_extract_spectral_variants()` (Step 4) — skipped for OLS/WLS
5. Calls `sw_unmix()` (Step 5)

### Simpler unmixing with OLS/WLS

If you don't need per-cell optimization, you can use a simpler method.
Steps 3–4 are skipped automatically:

```{r pipeline-ols}
result <- sw_unmix_pipeline(
  control_dir  = file.path(base_dir, "Raw", "Set1", "Reference Group"),
  sample_input = file.path(base_dir, "Raw", "Set1", "Stained"),
  cytometer    = "aurora",
  control_file = file.path(base_dir, "fcs_control_file.csv"),
  method       = "OLS",
  output_dir   = "unmix_output"
)
# No unstained_fcs needed — AF extraction is skipped
```

## Connecting to the downstream pipeline

After unmixing, the output FCS files can flow directly into the rest of
the SpectraWeaveR pipeline (gating → QC → batch correction → clustering):

```{r downstream}
# Load the unmixed files
fs <- sw_load_unmixed(result$unmixed$output_dir)

# Remove margin events (must be done BEFORE transformation)
ff_list <- lapply(seq_along(fs), function(i) {
  sw_remove_margins(fs[[i]])
})
names(ff_list) <- flowCore::sampleNames(fs)

# Continue with the rest of the pipeline...
# sw_signal_qc_batch(ff_list, ...)
# sw_batch_correct(...)
# sw_cluster(...)
```

Or use `run_pipeline()` with `unmix_from_raw = TRUE` to run everything
from raw spectral data through to clustering in a single call:

```{r full-pipeline}
results <- run_pipeline(
  fcs_dir         = file.path(base_dir, "Raw", "Set1", "Stained"),
  sample_meta     = sample_meta,
  markers         = markers,
  lineage_markers = lineage_markers,
  unmix_from_raw  = TRUE,
  control_dir     = file.path(base_dir, "Raw", "Set1", "Reference Group"),
  unstained_fcs   = file.path(base_dir, "Raw", "Set1", "Unstained",
                              "D1 Spleen_Set1.fcs"),
  cytometer       = "aurora",
  unmix_method    = "AutoSpectral"
)
```

## Multi-tissue unmixing

The example dataset demonstrates a common multi-tissue scenario: spleen,
lung, and liver samples each have distinct autofluorescence profiles.
The step-by-step workflow handles this by extracting tissue-specific AF
and matching it during unmixing.

```{r multi-tissue}
# Steps 1-2: Same for all tissues (controls are shared)
control_dir <- file.path(base_dir, "Raw", "Set1", "Reference Group")
control_file <- file.path(base_dir, "fcs_control_file.csv")

setup <- sw_autospectral_setup(
  control_dir  = control_dir,
  cytometer    = "aurora",
  control_file = control_file
)
controls <- sw_prepare_controls(setup)

# Step 3: Extract tissue-specific AF from each unstained sample
unstained_dir <- file.path(base_dir, "Raw", "Set1", "Unstained")
af_spectra <- sw_extract_af_spectra(
  unstained_fcs = list(
    spleen = file.path(unstained_dir, "D1 Spleen_Set1.fcs"),
    lung   = file.path(unstained_dir, "D2 Lung_Set1.fcs"),
    liver  = file.path(unstained_dir, "D3 Liver_Set1.fcs")
  ),
  setup   = setup,
  spectra = controls$spectra,
  refine  = TRUE
)

# Step 4: Extract variants using spleen AF (controls are from spleen)
variants <- sw_extract_spectral_variants(
  setup      = setup,
  spectra    = controls$spectra,
  af_spectra = af_spectra$spleen
)

# Step 5: Unmix each tissue with its matched AF spectra
stained_dir <- file.path(base_dir, "Raw", "Set1", "Stained")

spleen_result <- sw_unmix(
  input        = file.path(stained_dir, "D4 Spleen_Set1.fcs"),
  spectra      = controls$spectra,
  setup        = setup,
  flow_control = controls$flow_control,
  af_spectra   = af_spectra$spleen,
  spectra_variants = variants
)

lung_result <- sw_unmix(
  input        = file.path(stained_dir, "D5 Lung_Set1.fcs"),
  spectra      = controls$spectra,
  setup        = setup,
  flow_control = controls$flow_control,
  af_spectra   = af_spectra$lung,
  spectra_variants = variants
)

liver_result <- sw_unmix(
  input        = file.path(stained_dir, "D6 Liver_Set1.fcs"),
  spectra      = controls$spectra,
  setup        = setup,
  flow_control = controls$flow_control,
  af_spectra   = af_spectra$liver,
  spectra_variants = variants
)
```

## Tips and best practices

### Performance

- **Install AutoSpectralRcpp** for a significant speed improvement in
  AF extraction and unmixing.
- Use `parallel = TRUE` and set `threads` to the number of available
  cores for large experiments.
- The `speed` parameter in `sw_unmix()` controls the precision-speed
  trade-off: `"fast"` tests 1 variant per cell, `"medium"` tests 3, and
  `"slow"` tests 10.
- Adjust `chunk_size` (default `2e6`) for memory management on
  constrained systems.

### Quality control

- Always review the auto-generated control file before proceeding from
  Step 1 to Step 2.
- Check the QC figures generated in the output directory — they show
  gating, spectral signatures, and similarity heatmaps.
- Use `refine = TRUE` (the default) for tissue samples with complex
  autofluorescence — the second-pass refinement improves AF modeling.

### Reproducibility

- All intermediate objects are preserved in the return values for full
  provenance tracking.
- The `sw_unmix_pipeline()` orchestrator stores the complete chain of
  results for reproducibility.
- Save intermediate results (e.g., `spectra`, `af_spectra`, `variants`)
  as RDS files for reuse across sessions:

```{r save-intermediates}
saveRDS(controls$spectra, "spectra.rds")
saveRDS(af_spectra, "af_spectra.rds")
saveRDS(variants, "spectral_variants.rds")

# Reload in a new session
spectra  <- readRDS("spectra.rds")
af_specs <- readRDS("af_spectra.rds")
vars     <- readRDS("spectral_variants.rds")
```

## References

- **AutoSpectral documentation:**
  [Full Workflow Example](https://drcytometer.github.io/AutoSpectral/articles/01_Full_AutoSpectral_Workflow.html)
  — the original AutoSpectral tutorial using this dataset.
- **AutoSpectral blog post:**
  [Full Workflow Example](https://www.colibri-cytometry.com/post/autospectral-full-workflow-example)
  on Colibri Cytometry.
- **Example data:**
  [Mendeley Data (ch5dnspd79)](https://data.mendeley.com/datasets/ch5dnspd79/1)
  — 9-colour Aurora panel with spleen, lung, and liver samples.
- **AutoSpectral GitHub:**
  [DrCytometer/AutoSpectral](https://github.com/DrCytometer/AutoSpectral)

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
