---
title: "Tutorial Part 4: Clustering & Visualization"
vignette: >
  %\VignetteIndexEntry{Tutorial Part 4: Clustering & Visualization}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This is **Part 4** of the SpectraWeaveR end-to-end tutorial series:

1. [Data Loading & Preprocessing](tutorial-01-loading-preprocessing.qmd)
2. [Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)
3. [Batch Correction](tutorial-03-batch-correction.qmd)
4. **Clustering & Visualization** (this vignette)
5. [Pipeline Frameworks](tutorial-05-pipeline-frameworks.qmd)

This part covers: clustering with kohonen SOM, extracting assignments,
computing MFIs, heatmap visualization, predicting clusters on new data,
dimensionality reduction (PCA and UMAP), and unmixing quality diagnostics.

## Synthetic data setup

We create a batch-corrected tibble that represents the output of Part 3.

```{r}
#| label: synthetic-data
library(SpectraWeaveR)

set.seed(42)

markers <- c("CD3", "CD4", "CD8", "CD19", "CD56", "HLADR")
lineage_markers <- c("CD3", "CD4", "CD8", "CD19", "CD56")

# Simulate 3 populations of batch-corrected data
n_per_pop <- 200
pop1 <- data.frame(
  CD3 = rnorm(n_per_pop, 2.0, 0.3), CD4 = rnorm(n_per_pop, 2.5, 0.4),
  CD8 = rnorm(n_per_pop, 0.1, 0.2), CD19 = rnorm(n_per_pop, 0.05, 0.1),
  CD56 = rnorm(n_per_pop, 0.05, 0.1), HLADR = rnorm(n_per_pop, 0.3, 0.2)
)
pop2 <- data.frame(
  CD3 = rnorm(n_per_pop, 1.8, 0.3), CD4 = rnorm(n_per_pop, 0.1, 0.2),
  CD8 = rnorm(n_per_pop, 2.2, 0.4), CD19 = rnorm(n_per_pop, 0.05, 0.1),
  CD56 = rnorm(n_per_pop, 0.1, 0.1), HLADR = rnorm(n_per_pop, 0.2, 0.15)
)
pop3 <- data.frame(
  CD3 = rnorm(n_per_pop, 0.05, 0.1), CD4 = rnorm(n_per_pop, 0.05, 0.1),
  CD8 = rnorm(n_per_pop, 0.05, 0.1), CD19 = rnorm(n_per_pop, 2.0, 0.3),
  CD56 = rnorm(n_per_pop, 1.5, 0.3), HLADR = rnorm(n_per_pop, 1.2, 0.3)
)

corrected <- tibble::as_tibble(rbind(pop1, pop2, pop3))
corrected$sample <- rep(paste0("S", 1:6), each = 100)
corrected$batch  <- rep(rep(c("B1", "B2"), each = 3), each = 100)
corrected$condition <- rep(rep(c("control", "treatment"), 3), each = 100)

cat("Corrected data:", nrow(corrected), "cells,",
    length(markers), "markers\n")
```

## Clustering with kohonen SOM

`sw_cluster_run()` performs Self-Organizing Map (SOM) clustering followed by
hierarchical metaclustering. It also supports FastPG graph-based
clustering when that package is available.

```{r}
#| label: cluster
cluster_result <- sw_cluster_run(
  corrected,
  lineage_markers = lineage_markers,
  method = "som",
  xdim = 5, ydim = 5,
  n_metaclusters = 5,
  seed = 42
)

stopifnot(inherits(cluster_result, "sw_cluster_result"))
stopifnot(is.integer(cluster_result$assignments))
stopifnot(length(cluster_result$assignments) == nrow(corrected))
stopifnot(cluster_result$method == "som")
stopifnot(cluster_result$n_clusters == 5)
cat("Clustering complete:", cluster_result$n_clusters, "metaclusters\n")
```

### Extract cluster assignments

```{r}
#| label: get-assignments
assignments <- sw_cluster_assignments(cluster_result)
stopifnot(is.integer(assignments))
stopifnot(length(assignments) == nrow(corrected))
cat("Cluster distribution:\n")
print(table(assignments))
```

### Compute median fluorescence intensities (MFI)

```{r}
#| label: cluster-mfis
mfis <- sw_cluster_mfi(cluster_result)
stopifnot(is.data.frame(mfis))
stopifnot("cluster" %in% names(mfis))
stopifnot(nrow(mfis) == cluster_result$n_clusters)
stopifnot(all(lineage_markers %in% names(mfis)))
cat("MFI table:\n")
print(mfis)
```

### Cluster heatmap

`sw_plot_clusters()` generates a heatmap of median marker expression
per cluster and saves it to a PDF file.

```{r}
#| label: plot-clusters
heatmap_file <- tempfile(fileext = ".pdf")
result_path <- sw_plot_clusters(cluster_result, plot_file = heatmap_file)
stopifnot(file.exists(heatmap_file))
cat("Heatmap saved to:", heatmap_file, "\n")
cat("File size:", file.size(heatmap_file), "bytes\n")
```

### Predict clusters for new data

`sw_cluster_predict()` maps new cells onto the trained SOM and assigns
them to metaclusters.

```{r}
#| label: predict-clusters
# Create new data (simulate 50 "new" cells)
new_data <- tibble::tibble(
  CD3 = rnorm(50, 2.0, 0.3), CD4 = rnorm(50, 2.5, 0.4),
  CD8 = rnorm(50, 0.1, 0.2), CD19 = rnorm(50, 0.05, 0.1),
  CD56 = rnorm(50, 0.05, 0.1)
)

predicted <- sw_cluster_predict(cluster_result, newdata = new_data)
stopifnot(inherits(predicted, "sw_cluster_result"))
stopifnot(length(predicted$assignments) == 50)
cat("Predicted clusters for 50 new cells:\n")
print(table(predicted$assignments))
```

## Dimensionality reduction

### PCA

`sw_dimred_run()` computes a 2D embedding. PCA is always available
(no extra dependencies). The result includes a `variance_explained`
attribute.

```{r}
#| label: dimred-pca
pca_result <- sw_dimred_run(
  corrected, markers = markers,
  method = "pca", seed = 42
)

stopifnot(is.data.frame(pca_result))
stopifnot(all(c("dim1", "dim2") %in% names(pca_result)))
stopifnot(attr(pca_result, "method") == "pca")
var_expl <- attr(pca_result, "variance_explained")
stopifnot(!is.null(var_expl))
stopifnot(length(var_expl) == 2)
cat("PCA variance explained:", round(var_expl * 100, 1), "%\n")
head(pca_result[, c("dim1", "dim2", "sample", "batch")])
```

### UMAP (optional)

UMAP requires the `uwot` package.

```{r}
#| label: dimred-umap
#| eval: !expr requireNamespace("uwot", quietly = TRUE)
umap_result <- sw_dimred_run(
  corrected, markers = markers,
  method = "umap", n_neighbors = 15, min_dist = 0.2, seed = 42
)

stopifnot(is.data.frame(umap_result))
stopifnot(all(c("dim1", "dim2") %in% names(umap_result)))
stopifnot(attr(umap_result, "method") == "umap")
cat("UMAP complete:", nrow(umap_result), "cells\n")
```

### Plot dimensionality reduction

`sw_plot_dimred()` creates a scatter plot colored by any column.

```{r}
#| label: plot-dimred
#| fig-width: 10
#| fig-height: 4
#| eval: !expr requireNamespace("ggplot2", quietly = TRUE)
# Add cluster labels for visualization
pca_result$cluster <- factor(assignments)

p_batch <- sw_plot_dimred(pca_result, color_by = "batch",
                           title = "PCA colored by batch")
stopifnot(inherits(p_batch, "gg"))

p_cluster <- sw_plot_dimred(pca_result, color_by = "cluster",
                             title = "PCA colored by cluster")
stopifnot(inherits(p_cluster, "gg"))

cat("Created 2 dimred plots\n")
print(p_batch)
```

## Unmixing quality diagnostics

These functions assess the quality of spectral unmixing by examining
spillover spreading and per-channel statistics. They can be applied
post-hoc to any unmixed data.

### Spillover spreading matrix (SSM)

`sw_unmix_spillover_matrix()` quantifies cross-talk between
fluorophores based on reference spectra. Here we create synthetic
reference spectra to demonstrate.

```{r}
#| label: ssm
# Create synthetic reference spectra (6 fluorophores, 10 detectors)
# Each fluorophore peaks in a different detector with some cross-talk
spectra <- matrix(0, nrow = 6, ncol = 10,
                  dimnames = list(
                    c("BV421", "BV510", "BV605", "BV711", "PE", "APC"),
                    paste0("Det", 1:10)
                  ))
# Each row: primary peak + minor cross-talk
spectra[1, ] <- c(1.0, 0.3, 0.05, 0, 0, 0, 0, 0, 0, 0)
spectra[2, ] <- c(0.1, 0.8, 0.4, 0.1, 0, 0, 0, 0, 0, 0)
spectra[3, ] <- c(0, 0.05, 0.2, 1.0, 0.3, 0.05, 0, 0, 0, 0)
spectra[4, ] <- c(0, 0, 0, 0.1, 0.3, 1.0, 0.4, 0.1, 0, 0)
spectra[5, ] <- c(0, 0, 0, 0, 0, 0.1, 0.3, 1.0, 0.3, 0)
spectra[6, ] <- c(0, 0, 0, 0, 0, 0, 0, 0.1, 0.4, 1.0)

ssm <- sw_unmix_spillover_matrix(spectra)
stopifnot(inherits(ssm, "sw_ssm"))
stopifnot(is.matrix(ssm$matrix))
stopifnot(nrow(ssm$matrix) == 6)
stopifnot(all(diag(ssm$matrix) == 0))  # No self-spreading
stopifnot(all(ssm$matrix >= 0))
stopifnot(all(ssm$matrix <= 1 + 1e-10))

cat("SSM summary:\n")
print(ssm$summary)
```

### SSM with empirical data

When an unmixed flowFrame is provided, the SSM combines spectra-based
and empirical (correlation-based) estimates.

```{r}
#| label: ssm-empirical
# Use marker names matching the spectra rownames
mat_unmixed <- matrix(
  abs(rnorm(600, 5000, 2000)),
  ncol = 6,
  dimnames = list(NULL, c("BV421", "BV510", "BV605", "BV711", "PE", "APC"))
)
ff_unmixed <- flowCore::flowFrame(mat_unmixed)

ssm_emp <- sw_unmix_spillover_matrix(spectra, unmixed_ff = ff_unmixed)
stopifnot(inherits(ssm_emp, "sw_ssm"))
cat("Empirical SSM computed (combined spectra + data)\n")
```

### Plot SSM heatmap

```{r}
#| label: plot-ssm
ssm_pdf <- tempfile(fileext = ".pdf")
sw_plot_spillover_matrix(ssm, plot_file = ssm_pdf)
stopifnot(file.exists(ssm_pdf))
cat("SSM heatmap saved to:", ssm_pdf, "\n")
```

### Unmixing quality per channel

`sw_unmix_quality()` computes mean, SD, CV, and percentage of negative
values per channel — flagging channels with high CV.

```{r}
#| label: unmixing-quality
quality <- sw_unmix_quality(
  ff_unmixed,
  channels = c("BV421", "BV510", "BV605", "BV711", "PE", "APC"),
  cv_threshold = 0.5
)

stopifnot(is.data.frame(quality))
stopifnot(all(c("channel", "mean", "sd", "cv", "pct_negative", "flagged") %in%
              names(quality)))
stopifnot(nrow(quality) == 6)
cat("Per-channel quality:\n")
print(quality)
```

## Clean up

```{r}
#| label: cleanup
unlink(heatmap_file)
unlink(ssm_pdf)
cat("Tutorial Part 4 complete. All functions verified.\n")
```

## Functions covered

| Function | Module |
|----------|--------|
| `sw_cluster_run()` | cluster |
| `sw_cluster_assignments()` | cluster |
| `sw_cluster_mfi()` | cluster |
| `sw_plot_clusters()` | cluster |
| `sw_cluster_predict()` | cluster |
| `sw_dimred_run()` | dimred |
| `sw_plot_dimred()` | dimred |
| `sw_unmix_spillover_matrix()` | unmix_diagnostics |
| `sw_plot_spillover_matrix()` | unmix_diagnostics |
| `sw_unmix_quality()` | unmix_diagnostics |

**Previous:** [Part 3 — Batch Correction](tutorial-03-batch-correction.qmd)\
**Next:** [Part 5 — Pipeline Frameworks](tutorial-05-pipeline-frameworks.qmd)
