---
title: "Tutorial Part 3: Batch Correction"
vignette: >
  %\VignetteIndexEntry{Tutorial Part 3: Batch Correction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This is **Part 3** of the SpectraWeaveR end-to-end tutorial series:

1. [Data Loading & Preprocessing](tutorial-01-loading-preprocessing.qmd)
2. [Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)
3. **Batch Correction** (this vignette)
4. [Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
5. [Pipeline Frameworks](tutorial-05-pipeline-frameworks.qmd)

This part covers: preparing data for batch correction, the modular
workflow (normalize → SOM → correct), the all-in-one wrapper, and all
diagnostic/evaluation functions.

## Synthetic data setup

We recreate synthetic data and run preprocessing (Parts 1–2) to get
QC-cleaned flowFrames ready for batch correction.

```{r}
#| label: synthetic-data
library(SpectraWeaveR)
library(flowCore)

set.seed(42)

make_tutorial_fcs <- function(n_per_pop = 150, batch_shift = 0) {
  pop_cd4 <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 150000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 120000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  50000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  40000,  9000)),
    CD3  = abs(rnorm(n_per_pop, 3000 + batch_shift, 500)),
    CD4  = abs(rnorm(n_per_pop, 5000 + batch_shift, 800)),
    CD8  = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    CD19 = abs(rnorm(n_per_pop,  300 + batch_shift, 100)),
    CD56 = abs(rnorm(n_per_pop,  200 + batch_shift, 100)),
    HLADR = abs(rnorm(n_per_pop,  500 + batch_shift, 200))
  )
  pop_cd8 <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 140000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 110000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  45000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  36000,  9000)),
    CD3  = abs(rnorm(n_per_pop, 2500 + batch_shift, 500)),
    CD4  = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    CD8  = abs(rnorm(n_per_pop, 4500 + batch_shift, 700)),
    CD19 = abs(rnorm(n_per_pop,  350 + batch_shift, 120)),
    CD56 = abs(rnorm(n_per_pop,  250 + batch_shift, 100)),
    HLADR = abs(rnorm(n_per_pop,  450 + batch_shift, 180))
  )
  pop_bcell <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 130000, 18000)),
    `FSC-H` = abs(rnorm(n_per_pop, 100000, 16000)),
    `SSC-A` = abs(rnorm(n_per_pop,  40000,  8000)),
    `SSC-H` = abs(rnorm(n_per_pop,  32000,  7000)),
    CD3  = abs(rnorm(n_per_pop,  300 + batch_shift, 120)),
    CD4  = abs(rnorm(n_per_pop,  350 + batch_shift, 130)),
    CD8  = abs(rnorm(n_per_pop,  280 + batch_shift, 100)),
    CD19 = abs(rnorm(n_per_pop, 4000 + batch_shift, 600)),
    CD56 = abs(rnorm(n_per_pop, 3500 + batch_shift, 500)),
    HLADR = abs(rnorm(n_per_pop, 2000 + batch_shift, 400))
  )
  mat <- rbind(pop_cd4, pop_cd8, pop_bcell)
  mat <- mat[sample(nrow(mat)), ]
  storage.mode(mat) <- "double"
  flowFrame(mat)
}

# Build 6 QC-cleaned flowFrames (3 per batch with batch shift)
sample_meta <- tibble::tibble(
  sample    = paste0("S", 1:6),
  batch     = rep(c("B1", "B2"), each = 3),
  condition = rep(c("control", "treatment"), 3)
)

ff_list <- list()
for (i in seq_len(nrow(sample_meta))) {
  shift <- if (sample_meta$batch[i] == "B2") 300 else 0
  ff_list[[sample_meta$sample[i]]] <- make_tutorial_fcs(
    n_per_pop = 150, batch_shift = shift
  )
}

markers <- c("CD3", "CD4", "CD8", "CD19", "CD56", "HLADR")
cat("Created", length(ff_list), "samples with", length(markers), "markers\n")
```

## Prepare data for batch correction

`sw_correct_prepare()` converts a list of flowFrames into a
single arcsinh-transformed tibble with metadata columns. The default
cofactor of 6000 is appropriate for spectral flow cytometry.

```{r}
#| label: prepare
uncorrected <- sw_correct_prepare(
  ff_list,
  sample_meta = sample_meta,
  markers = markers,
  cofactor = 6000
)

stopifnot(is.data.frame(uncorrected))
stopifnot(all(markers %in% names(uncorrected)))
stopifnot(all(c("sample", "batch", "condition", ".cell_id") %in%
              names(uncorrected)))
cat("Prepared", nrow(uncorrected), "cells for correction\n")
cat("Batches:", paste(unique(uncorrected$batch), collapse = ", "), "\n")
head(uncorrected[, c(markers[1:3], "sample", "batch")])
```

## Detect batch effects

`sw_correct_detect_batch()` produces diagnostic plots and metrics. It
delegates to `cyCombine::detect_batch_effect_express()`.

```{r}
#| label: detect-batch
detect_result <- tryCatch({
  res <- sw_correct_detect_batch(uncorrected, markers = markers, seed = 42)
  cat("Batch effect detection complete\n")
  res
}, error = function(e) {
  cat("detect_batch_effect note:", conditionMessage(e), "\n")
  NULL
})
```

## Modular batch correction workflow

The modular workflow gives fine-grained control over each step:

1. **Normalize** — per-batch z-score or rank normalization
2. **Create SOM** — Self-Organizing Map clusters for ComBat
3. **Correct data** — ComBat batch correction within each cluster

### Step 1: Normalize

```{r}
#| label: normalize
normalized <- sw_correct_normalize(uncorrected, markers = markers,
                            norm_method = "scale")
stopifnot(is.data.frame(normalized))
stopifnot(nrow(normalized) == nrow(uncorrected))
stopifnot(all(markers %in% names(normalized)))
cat("Normalized", nrow(normalized), "cells\n")
```

### Step 2: Create SOM clusters

```{r}
#| label: create-som
label <- sw_correct_som(normalized, markers = markers,
                        xdim = 4, ydim = 4, rlen = 10, seed = 42)
stopifnot(is.numeric(label) || is.integer(label))
stopifnot(length(label) == nrow(uncorrected))
cat("SOM clusters:", length(unique(label)), "nodes\n")
```

### Step 3: Apply ComBat correction

```{r}
#| label: correct-data
corrected_modular <- sw_correct_apply(
  uncorrected,
  label = label,
  markers = markers,
  covar = "condition"
)

stopifnot(is.data.frame(corrected_modular))
stopifnot(nrow(corrected_modular) == nrow(uncorrected))
stopifnot(all(markers %in% names(corrected_modular)))
cat("Modular correction complete:", nrow(corrected_modular), "cells\n")
```

## All-in-one batch correction

`sw_correct_run()` combines normalize → SOM → correct into a single call.

```{r}
#| label: batch-correct-allinone
corrected <- sw_correct_run(
  uncorrected,
  markers = markers,
  covar = "condition",
  xdim = 4, ydim = 4, rlen = 10,
  seed = 42
)

stopifnot(is.data.frame(corrected))
stopifnot(nrow(corrected) == nrow(uncorrected))
stopifnot(all(markers %in% names(corrected)))
cat("All-in-one correction complete:", nrow(corrected), "cells\n")
```

## Evaluate correction quality

### Quick MAD-based evaluation

`sw_correct_evaluate_quick()` compares batch-median MAD before and after
correction.

```{r}
#| label: evaluate-correction
eval_result <- sw_correct_evaluate_quick(uncorrected, corrected, markers = markers)
stopifnot(is.list(eval_result))
stopifnot(all(c("emd", "mad", "improved", "emd_reduction_pct") %in%
              names(eval_result)))
cat("Correction improved:", eval_result$improved, "\n")
cat("MAD reduction:", eval_result$emd_reduction_pct, "%\n")
print(eval_result$mad)
```

### EMD-based evaluation

`sw_correct_emd()` computes Earth Mover's Distance between batch
distributions within each SOM cluster. `sw_correct_evaluate_emd()` compares
EMD before vs. after correction.

```{r}
#| label: emd-evaluation
# Add cluster labels to both datasets for EMD computation
uncorrected_labeled <- uncorrected
uncorrected_labeled$label <- label

corrected_labeled <- corrected
corrected_labeled$label <- label

# Compute EMD on labeled data
emd_scores <- sw_correct_emd(
  uncorrected_labeled,
  markers = markers,
  cell_col = "label",
  batch_col = "batch"
)
cat("EMD scores computed\n")

# Compare before vs after
emd_eval <- sw_correct_evaluate_emd(
  uncorrected_labeled,
  corrected_labeled,
  markers = markers,
  cell_col = "label"
)
cat("EMD evaluation complete\n")
```

### MAD-based evaluation (cyCombine)

```{r}
#| label: mad-evaluation
mad_eval <- sw_correct_evaluate_mad(
  uncorrected_labeled,
  corrected_labeled,
  markers = markers,
  cell_col = "label"
)
cat("MAD evaluation complete\n")
```

## Visualization

### Density plots by batch

`sw_plot_batch_densities()` creates ridgeline density plots comparing
marker distributions across batches before and after correction.

```{r}
#| label: density-plots
#| fig-width: 10
#| fig-height: 8
density_plot <- sw_plot_batch_densities(
  uncorrected, corrected,
  markers = markers[1:3]
)

if (inherits(density_plot, "gg") ||
    (is.list(density_plot) && length(density_plot) > 0)) {
  cat("Density plots created successfully\n")
}
```

### Dimensionality reduction colored by batch

`sw_plot_batch_dimred()` runs PCA or UMAP on the data and colors
points by batch — useful for visualizing batch effects.

```{r}
#| label: batch-dimred
#| fig-width: 8
#| fig-height: 6
dimred_plot <- sw_plot_batch_dimred(
  uncorrected, markers = markers,
  type = "pca", name = "Before correction"
)

if (inherits(dimred_plot, "gg")) {
  cat("Batch dimred plot created successfully\n")
  print(dimred_plot)
}
```

## Clean up

```{r}
#| label: cleanup
cat("Tutorial Part 3 complete. All functions verified.\n")
```

## Functions covered

| Function | Purpose |
|----------|---------|
| `sw_correct_prepare()` | flowFrame list → arcsinh tibble |
| `sw_correct_detect_batch()` | Diagnostic density/MDS/EMD plots |
| `sw_correct_normalize()` | Per-batch normalization |
| `sw_correct_som()` | SOM clustering for ComBat |
| `sw_correct_apply()` | ComBat correction with pre-computed labels |
| `sw_correct_run()` | All-in-one batch correction |
| `sw_correct_evaluate_quick()` | Quick MAD-based evaluation |
| `sw_correct_emd()` | Earth Mover's Distance per marker/cluster |
| `sw_correct_evaluate_emd()` | Compare EMD before vs. after |
| `sw_correct_evaluate_mad()` | Compare MAD before vs. after |
| `sw_plot_batch_densities()` | Ridgeline density plots |
| `sw_plot_batch_dimred()` | PCA/UMAP colored by batch |

**Previous:** [Part 2 — Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)\
**Next:** [Part 4 — Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
