---
title: "Tutorial Part 2: Gating & Signal QC"
vignette: >
  %\VignetteIndexEntry{Tutorial Part 2: Gating & Signal QC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This is **Part 2** of the SpectraWeaveR end-to-end tutorial series:

1. [Data Loading & Preprocessing](tutorial-01-loading-preprocessing.qmd)
2. **Gating & Signal QC** (this vignette)
3. [Batch Correction](tutorial-03-batch-correction.qmd)
4. [Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
5. [Pipeline Frameworks](tutorial-05-pipeline-frameworks.qmd)

This part covers: building gating templates, automated gating with
openCyto, extracting gated populations, scale transforms, and signal
quality control with PeacoQC.

## Synthetic data setup

We recreate the synthetic FCS data from Part 1 so this vignette is
self-contained.

```{r}
#| label: synthetic-data
library(SpectraWeaveR)
library(flowCore)

set.seed(42)

make_tutorial_fcs <- function(n_per_pop = 150, batch_shift = 0) {
  pop_cd4 <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 150000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 120000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  50000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  40000,  9000)),
    `BV421-A` = abs(rnorm(n_per_pop, 3000 + batch_shift, 500)),
    `BV510-A` = abs(rnorm(n_per_pop, 5000 + batch_shift, 800)),
    `BV605-A` = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    `BV711-A` = abs(rnorm(n_per_pop,  300 + batch_shift, 100)),
    `PE-A`    = abs(rnorm(n_per_pop,  200 + batch_shift, 100)),
    `APC-A`   = abs(rnorm(n_per_pop,  500 + batch_shift, 200))
  )
  pop_cd8 <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 140000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 110000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  45000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  36000,  9000)),
    `BV421-A` = abs(rnorm(n_per_pop, 2500 + batch_shift, 500)),
    `BV510-A` = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    `BV605-A` = abs(rnorm(n_per_pop, 4500 + batch_shift, 700)),
    `BV711-A` = abs(rnorm(n_per_pop,  350 + batch_shift, 120)),
    `PE-A`    = abs(rnorm(n_per_pop,  250 + batch_shift, 100)),
    `APC-A`   = abs(rnorm(n_per_pop,  450 + batch_shift, 180))
  )
  pop_bcell <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 130000, 18000)),
    `FSC-H` = abs(rnorm(n_per_pop, 100000, 16000)),
    `SSC-A` = abs(rnorm(n_per_pop,  40000,  8000)),
    `SSC-H` = abs(rnorm(n_per_pop,  32000,  7000)),
    `BV421-A` = abs(rnorm(n_per_pop,  300 + batch_shift, 120)),
    `BV510-A` = abs(rnorm(n_per_pop,  350 + batch_shift, 130)),
    `BV605-A` = abs(rnorm(n_per_pop,  280 + batch_shift, 100)),
    `BV711-A` = abs(rnorm(n_per_pop, 4000 + batch_shift, 600)),
    `PE-A`    = abs(rnorm(n_per_pop, 3500 + batch_shift, 500)),
    `APC-A`   = abs(rnorm(n_per_pop, 2000 + batch_shift, 400))
  )
  mat <- rbind(pop_cd4, pop_cd8, pop_bcell)
  mat <- mat[sample(nrow(mat)), ]
  storage.mode(mat) <- "double"
  flowFrame(mat)
}

# Write 6 FCS files
fcs_dir <- file.path(tempdir(), "tutorial_fcs_02")
dir.create(fcs_dir, showWarnings = FALSE, recursive = TRUE)

sample_info <- data.frame(
  file   = paste0("sample", 1:6, ".fcs"),
  sample = paste0("S", 1:6),
  batch  = rep(c("B1", "B2"), each = 3),
  condition = rep(c("control", "treatment"), 3),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(sample_info))) {
  shift <- if (sample_info$batch[i] == "B2") 300 else 0
  ff <- make_tutorial_fcs(n_per_pop = 150, batch_shift = shift)
  write.FCS(ff, file.path(fcs_dir, sample_info$file[i]))
}

# Load data and run margin removal (from Part 1)
fs <- sw_load_unmixed(fcs_dir)
ff_list <- lapply(seq_along(fs), function(i) {
  sw_remove_margins(fs[[i]])
})
names(ff_list) <- sampleNames(fs)
cat("Loaded and margin-cleaned", length(ff_list), "samples\n")
```

## Build a gating template

`sw_build_gating_template()` writes a CSV gating template compatible
with the openCyto framework. Five pre-built templates are available:
`"lymphocyte"`, `"myeloid"`, `"nk"`, `"treg"`, `"full_pbmc"`.

```{r}
#| label: build-template
template_file <- file.path(tempdir(), "lymphocyte_template.csv")
sw_build_gating_template(template_file, template_type = "lymphocyte")

stopifnot(file.exists(template_file))
template_df <- read.csv(template_file)
cat("Template gates:\n")
print(template_df[, c("alias", "pop", "parent", "dims", "gating_method")])
stopifnot(nrow(template_df) >= 1)
```

## Automated gating

`sw_gate()` loads FCS files, creates a GatingSet, optionally transforms
channels, and applies the gating template. Because synthetic data may not
satisfy the density assumptions of all gating algorithms, we wrap this in
a `tryCatch`.

```{r}
#| label: automated-gating
fcs_paths <- file.path(fcs_dir, sample_info$file[1:2])

gating_result <- tryCatch({
  gs <- sw_gate(fcs_paths, gating_template = template_file)
  stopifnot(is(gs, "GatingSet"))
  cat("Gating succeeded. GatingSet with",
      length(gs), "samples\n")
  cat("Gate nodes:",
      paste(flowWorkspace::gs_get_pop_paths(gs), collapse = ", "), "\n")
  gs
}, error = function(e) {
  cat("Gating on synthetic data raised an expected error:\n ",
      conditionMessage(e), "\n")
  cat("(This is normal — real spectral data has clearer density features.)\n")
  NULL
})
```

### Extract gated populations

If gating succeeded, `sw_extract_gated()` extracts flowFrames from a
specific gate node.

```{r}
#| label: extract-gated
if (!is.null(gating_result)) {
  nodes <- flowWorkspace::gs_get_pop_paths(gating_result)
  # Use the deepest node available
  target_node <- nodes[length(nodes)]
  gated_list <- sw_extract_gated(gating_result, node = target_node)
  stopifnot(is.list(gated_list))
  stopifnot(length(gated_list) > 0)
  cat("Extracted", length(gated_list), "samples from node:", target_node, "\n")
  for (nm in names(gated_list)) {
    stopifnot(is(gated_list[[nm]], "flowFrame"))
  }
} else {
  cat("Skipping sw_extract_gated() — gating did not succeed on synthetic data.\n")
  cat("See the spectral-unmixing vignette for real-data examples.\n")
}
```

## Scale transforms

`sw_estimate_scale_transforms()` estimates per-channel logicle (or linear)
scale transformations. `sw_apply_scale_transforms()` applies them.

```{r}
#| label: scale-transforms
ff_test <- ff_list[[1]]

trans_result <- tryCatch({
  trans_list <- sw_estimate_scale_transforms(
    ff_test,
    fluo_method = "estimateLogicle",
    scatter_method = "none"
  )
  stopifnot(is(trans_list, "transformList"))
  cat("Estimated logicle transforms for",
      length(trans_list@transforms), "channels\n")

  # Apply the transforms
  ff_transformed <- sw_apply_scale_transforms(ff_test, trans_list)
  stopifnot(is(ff_transformed, "flowFrame"))
  stopifnot(nrow(ff_transformed) == nrow(ff_test))

  # Verify values changed
  orig_vals <- exprs(ff_test)[1:3, sw_get_fluor_channels(ff_test)[1]]
  new_vals  <- exprs(ff_transformed)[1:3, sw_get_fluor_channels(ff_test)[1]]
  stopifnot(!all(orig_vals == new_vals))
  cat("Transform applied successfully — values differ from original\n")

  trans_list
}, error = function(e) {
  cat("Logicle estimation note:", conditionMessage(e), "\n")
  cat("(estimateLogicle may struggle with purely positive synthetic data.)\n")
  NULL
})
```

## Signal quality control (PeacoQC)

PeacoQC uses Isolation Trees and Median Absolute Deviation (MAD) to
identify aberrant signal events — cells where instrument instability
caused anomalous readings.

### Single-sample QC

```{r}
#| label: signal-qc-single
ff_for_qc <- ff_list[[1]]
fluor_channels <- sw_get_fluor_channels(ff_for_qc)

qc_result <- sw_signal_qc(
  ff_for_qc,
  channels = fluor_channels,
  IT_limit = 0.55,
  MAD = 6
)

stopifnot(is.list(qc_result))
stopifnot("FinalFF" %in% names(qc_result))
stopifnot("GoodCells" %in% names(qc_result))
stopifnot("n_removed" %in% names(qc_result))
stopifnot("pct_removed" %in% names(qc_result))
stopifnot(is(qc_result$FinalFF, "flowFrame"))
stopifnot(nrow(qc_result$FinalFF) <= nrow(ff_for_qc))
cat("QC result: removed", qc_result$n_removed, "events",
    "(", qc_result$pct_removed, "%)\n")
```

### Batch QC across all samples

`sw_signal_qc_batch()` applies QC to a list of flowFrames and returns
a summary.

```{r}
#| label: signal-qc-batch
qc_batch <- sw_signal_qc_batch(
  ff_list,
  channels = fluor_channels,
  IT_limit = 0.55,
  MAD = 6
)

stopifnot(is.list(qc_batch))
stopifnot("cleaned" %in% names(qc_batch))
stopifnot("summary" %in% names(qc_batch))
stopifnot(is.list(qc_batch$cleaned))
stopifnot(length(qc_batch$cleaned) == length(ff_list))
stopifnot(is.data.frame(qc_batch$summary))

cat("Batch QC summary:\n")
print(qc_batch$summary)

# All cleaned frames should be flowFrames
for (nm in names(qc_batch$cleaned)) {
  stopifnot(is(qc_batch$cleaned[[nm]], "flowFrame"))
}
```

### QC summary with flagging

`sw_qc_summary()` flags samples that exceed a removal threshold (default 30%).

```{r}
#| label: qc-summary
qc_summary <- sw_qc_summary(qc_batch, threshold = 30)
stopifnot(is.data.frame(qc_summary))
stopifnot("flagged" %in% names(qc_summary))
stopifnot(is.logical(qc_summary$flagged))
cat("Flagged samples (>30% removal):", sum(qc_summary$flagged), "\n")
print(qc_summary[, c("sample", "n_before", "n_after", "pct_removed", "flagged")])
```

### Track events through the pipeline

```{r}
#| label: events-audit
# Build an audit trail across the preprocessing steps
raw_events <- sum(vapply(seq_along(fs), function(i) nrow(fs[[i]]), integer(1)))
margin_events <- sum(vapply(ff_list, nrow, integer(1)))
qc_events <- sum(vapply(qc_batch$cleaned, nrow, integer(1)))

audit <- sw_collect_events_retained(list(
  raw_loaded    = raw_events,
  after_margins = margin_events,
  after_qc      = qc_events
))

stopifnot(is.data.frame(audit))
stopifnot(all(c("step", "n_events", "pct_of_initial") %in% names(audit)))
cat("Event audit trail:\n")
print(audit)
```

## Clean up

```{r}
#| label: cleanup
unlink(fcs_dir, recursive = TRUE)
unlink(template_file)
cat("Tutorial Part 2 complete. All functions verified.\n")
```

## Functions covered

| Function | Module |
|----------|--------|
| `sw_build_gating_template()` | gate |
| `sw_gate()` | gate |
| `sw_extract_gated()` | gate |
| `sw_estimate_scale_transforms()` | transforms |
| `sw_apply_scale_transforms()` | transforms |
| `sw_signal_qc()` | qc |
| `sw_signal_qc_batch()` | qc |
| `sw_qc_summary()` | qc |

**Previous:** [Part 1 — Data Loading & Preprocessing](tutorial-01-loading-preprocessing.qmd)\
**Next:** [Part 3 — Batch Correction](tutorial-03-batch-correction.qmd)
