---
title: "Tutorial Part 5: Pipeline Frameworks"
vignette: >
  %\VignetteIndexEntry{Tutorial Part 5: Pipeline Frameworks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This is **Part 5** of the SpectraWeaveR end-to-end tutorial series:

1. [Data Loading & Preprocessing](tutorial-01-loading-preprocessing.qmd)
2. [Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)
3. [Batch Correction](tutorial-03-batch-correction.qmd)
4. [Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
5. **Pipeline Frameworks** (this vignette)

This part covers: the `run_pipeline()` end-to-end orchestrator, the
S7-based composable pipeline framework (`sw_step`, `sw_pipeline`,
pipeline manipulation, `%>>%` operator), and the `targets` backend for
incremental pipeline execution.

## The `run_pipeline()` orchestrator

`run_pipeline()` executes all pipeline steps in validated order:
load FCS → remove margins → (optional gating) → signal QC →
batch correction → clustering.

```{r}
#| label: run-pipeline
library(SpectraWeaveR)
library(flowCore)

set.seed(42)

# Create synthetic FCS files
fcs_dir <- file.path(tempdir(), "tutorial_pipeline")
dir.create(fcs_dir, showWarnings = FALSE, recursive = TRUE)
out_dir <- file.path(tempdir(), "tutorial_pipeline_output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

make_fcs <- function(filename, n = 300, batch_shift = 0) {
  mat <- matrix(
    c(abs(rnorm(n, 150000, 30000)),
      abs(rnorm(n,  50000, 15000)),
      abs(rnorm(n, 1000 + batch_shift, 300)),
      abs(rnorm(n, 2000 + batch_shift, 500)),
      abs(rnorm(n,  500 + batch_shift, 200))),
    ncol = 5,
    dimnames = list(NULL, c("FSC-A", "SSC-A", "CD3", "CD4", "CD8"))
  )
  storage.mode(mat) <- "double"
  ff <- flowFrame(mat)
  write.FCS(ff, file.path(fcs_dir, filename))
}

make_fcs("sample1.fcs", batch_shift = 0)
make_fcs("sample2.fcs", batch_shift = 200)

sample_meta <- tibble::tibble(
  file = c("sample1.fcs", "sample2.fcs"),
  sample = c("S1", "S2"),
  batch = c("B1", "B2")
)

result <- run_pipeline(
  fcs_dir = fcs_dir,
  sample_meta = sample_meta,
  markers = c("CD3", "CD4", "CD8"),
  lineage_markers = c("CD3", "CD4", "CD8"),
  gating_template = NULL,
  output_dir = out_dir
)

stopifnot(is.list(result))
stopifnot("flowset" %in% names(result))
stopifnot("qc_results" %in% names(result))
stopifnot("corrected" %in% names(result))
stopifnot("cluster_result" %in% names(result))

cat("Pipeline results:\n")
cat("  flowset:", length(result$flowset), "samples\n")
if (!is.null(result$corrected)) {
  stopifnot(is.data.frame(result$corrected))
  stopifnot(all(c("CD3", "CD4", "CD8") %in% names(result$corrected)))
  cat("  corrected:", nrow(result$corrected), "cells\n")
}
if (!is.null(result$cluster_result)) {
  stopifnot(inherits(result$cluster_result, "sw_cluster_result"))
  cat("  clusters:", result$cluster_result$n_clusters, "metaclusters\n")
}

# Check additional outputs
if (!is.null(result$cluster_assignments)) {
  stopifnot(is.integer(result$cluster_assignments))
  cat("  assignments:", length(result$cluster_assignments), "cells\n")
}
if (!is.null(result$cluster_mfis)) {
  stopifnot(is.data.frame(result$cluster_mfis))
  cat("  MFI table:", nrow(result$cluster_mfis), "clusters\n")
}

unlink(fcs_dir, recursive = TRUE)
unlink(out_dir, recursive = TRUE)
```

## Composable pipeline framework (S7)

The composable pipeline framework provides an S7-based step-and-pipeline
abstraction inspired by CytoPipeline. It lets you define, compose,
manipulate, and execute pipelines programmatically.

### Create and execute a step

`sw_step()` wraps any R function as a processing step. `execute_step()`
runs it.

```{r}
#| label: step-basics
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# Custom step: double the input
step_double <- sw_step("double", function(x) x * 2)
stopifnot(step_double@name == "double")
stopifnot(is.function(step_double@FUN))

result <- execute_step(step_double, 5)
stopifnot(result == 10)
cat("execute_step(double, 5) =", result, "\n")

# Step with arguments
step_add <- sw_step("add_n", function(x, n) x + n, ARGS = list(n = 3))
stopifnot(execute_step(step_add, 7) == 10)
cat("execute_step(add_n, 7) =", execute_step(step_add, 7), "\n")
```

### Create a pipeline

`sw_pipeline()` creates an ordered collection of steps.

```{r}
#| label: pipeline-create
#| eval: !expr requireNamespace("S7", quietly = TRUE)
pip <- sw_pipeline("math_ops", steps = list(
  sw_step("double", function(x) x * 2),
  sw_step("add_one", function(x) x + 1),
  sw_step("square", function(x) x^2)
))

stopifnot(pip@name == "math_ops")
cat("Created pipeline:", pip@name, "\n")
```

### Pipeline metadata

```{r}
#| label: pipeline-metadata
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# Length
n <- sw_pipeline_length(pip)
stopifnot(n == 3)
cat("Pipeline length:", n, "\n")

# Step names
step_names <- sw_pipeline_step_names(pip)
stopifnot(identical(step_names, c("double", "add_one", "square")))
cat("Step names:", paste(step_names, collapse = " -> "), "\n")

# Show pipeline structure
sw_pipeline_show(pip)
```

### Plot pipeline flowchart

```{r}
#| label: plot-pipeline
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# Text flowchart
sw_plot_pipeline(pip, style = "text")

# Data frame output
step_df <- sw_plot_pipeline(pip, style = "data")
stopifnot(is.data.frame(step_df))
stopifnot(all(c("step_number", "name", "n_args") %in% names(step_df)))
cat("\nStep data frame:\n")
print(step_df)
```

### Pipeline manipulation

Pipelines are immutable — manipulation functions return new pipelines.

```{r}
#| label: pipeline-manipulation
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# Add a step
pip2 <- sw_pipeline_add(pip, sw_step("negate", function(x) -x))
stopifnot(sw_pipeline_length(pip2) == 4)
stopifnot("negate" %in% sw_pipeline_step_names(pip2))
cat("After add:", paste(sw_pipeline_step_names(pip2), collapse = " -> "), "\n")

# Add at a specific position
pip3 <- sw_pipeline_add(pip, sw_step("halve", function(x) x / 2), after = 1)
stopifnot(sw_pipeline_step_names(pip3)[2] == "halve")
cat("After insert:", paste(sw_pipeline_step_names(pip3), collapse = " -> "), "\n")

# Remove a step
pip4 <- sw_pipeline_remove(pip2, "negate")
stopifnot(sw_pipeline_length(pip4) == 3)
stopifnot(!("negate" %in% sw_pipeline_step_names(pip4)))
cat("After remove:", paste(sw_pipeline_step_names(pip4), collapse = " -> "), "\n")

# Replace a step
pip5 <- sw_pipeline_replace(
  pip, "add_one",
  sw_step("add_ten", function(x) x + 10)
)
stopifnot("add_ten" %in% sw_pipeline_step_names(pip5))
stopifnot(!("add_one" %in% sw_pipeline_step_names(pip5)))
cat("After replace:", paste(sw_pipeline_step_names(pip5), collapse = " -> "), "\n")

# Concatenate two pipelines
pip_a <- sw_pipeline("part_a", steps = list(
  sw_step("step_a", function(x) x + 100)
))
pip_b <- sw_pipeline("part_b", steps = list(
  sw_step("step_b", function(x) x * 2)
))
pip_combined <- sw_pipeline_concat(pip_a, pip_b, name = "combined")
stopifnot(sw_pipeline_length(pip_combined) == 2)
stopifnot(identical(sw_pipeline_step_names(pip_combined), c("step_a", "step_b")))
cat("Concatenated:", paste(sw_pipeline_step_names(pip_combined), collapse = " -> "), "\n")
```

### Execute a pipeline

`sw_pipeline_run()` threads data through each step sequentially.

```{r}
#| label: pipeline-run
#| eval: !expr requireNamespace("S7", quietly = TRUE)
run_result <- sw_pipeline_run(pip, input = 5, trace = TRUE)

stopifnot(is.list(run_result))
stopifnot("result" %in% names(run_result))
stopifnot("intermediates" %in% names(run_result))
stopifnot("steps_completed" %in% names(run_result))

# 5 * 2 = 10, 10 + 1 = 11, 11^2 = 121
stopifnot(run_result$result == 121)
stopifnot(run_result$steps_completed == 3)
stopifnot(length(run_result$intermediates) == 3)
stopifnot(run_result$intermediates$double == 10)
stopifnot(run_result$intermediates$add_one == 11)
stopifnot(run_result$intermediates$square == 121)
cat("Pipeline result:", run_result$result, "(expected 121)\n")
```

### The `%>>%` pipe operator

The `%>>%` operator composes steps and/or pipelines into new pipelines.

```{r}
#| label: pipe-operator
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# Step %>>% Step
pip_from_pipe <- sw_step("triple", function(x) x * 3) %>>%
                 sw_step("minus_one", function(x) x - 1)

stopifnot(sw_pipeline_length(pip_from_pipe) == 2)

pipe_result <- sw_pipeline_run(pip_from_pipe, input = 4, trace = FALSE)
# 4 * 3 = 12, 12 - 1 = 11
stopifnot(pipe_result$result == 11)
cat("%>>% result:", pipe_result$result, "(expected 11)\n")

# Pipeline %>>% Step
pip_extended <- pip_from_pipe %>>% sw_step("halve", function(x) x / 2)
ext_result <- sw_pipeline_run(pip_extended, input = 4, trace = FALSE)
# 4 * 3 = 12, 12 - 1 = 11, 11 / 2 = 5.5
stopifnot(ext_result$result == 5.5)
cat("Extended result:", ext_result$result, "(expected 5.5)\n")
```

### Convenience step constructors

SpectraWeaveR provides 8 pre-built step constructors that wrap the
core pipeline functions with sensible input/output type declarations.

```{r}
#| label: step-constructors
#| eval: !expr requireNamespace("S7", quietly = TRUE)
# These constructors create ProcessingStep objects ready for pipeline use
step_read <- sw_step_read_fcs()
stopifnot(step_read@name == "read_fcs")
cat("sw_step_read_fcs:", step_read@name, "\n")

step_margins <- sw_step_remove_margins()
stopifnot(step_margins@name == "remove_margins")
cat("sw_step_remove_margins:", step_margins@name, "\n")

step_qc <- sw_step_signal_qc(IT_limit = 0.55)
stopifnot(step_qc@name == "signal_qc")
stopifnot(length(step_qc@ARGS) > 0)
cat("sw_step_signal_qc:", step_qc@name, "(", length(step_qc@ARGS), "args)\n")

step_norm <- sw_step_normalize(markers = c("CD3", "CD4"),
                                norm_method = "scale")
stopifnot(step_norm@name == "normalize")
cat("sw_step_normalize:", step_norm@name, "\n")

step_som <- sw_step_create_som(markers = c("CD3", "CD4"),
                                xdim = 4, ydim = 4)
stopifnot(step_som@name == "create_som")
cat("sw_step_create_som:", step_som@name, "\n")

step_correct <- sw_step_correct_data(
  label = integer(0), markers = c("CD3", "CD4")
)
stopifnot(step_correct@name == "correct_data")
cat("sw_step_correct_data:", step_correct@name, "\n")

step_bc <- sw_step_batch_correct(markers = c("CD3", "CD4"))
stopifnot(step_bc@name == "batch_correct")
cat("sw_step_batch_correct:", step_bc@name, "\n")

step_clust <- sw_step_cluster(
  lineage_markers = c("CD3", "CD4"),
  method = "som", n_metaclusters = 5
)
stopifnot(step_clust@name == "cluster")
cat("sw_step_cluster:", step_clust@name, "\n")

cat("\nAll 8 convenience step constructors verified.\n")
```

## targets backend

The `targets` backend converts each `sw_step()` into a
`targets::tar_target()` so reruns are incremental — only steps whose
definition changed (and their downstream descendants) are recomputed.

```{r}
#| label: targets-backend
#| eval: !expr requireNamespace("targets", quietly = TRUE) && requireNamespace("S7", quietly = TRUE)
# Create a simple pipeline for targets
pip_targets <- sw_pipeline("targets_demo", steps = list(
  sw_step("times3", function(x) x * 3),
  sw_step("plus5",  function(x) x + 5),
  sw_step("sqrt",   function(x) sqrt(x))
))

# First: write the _targets.R script without executing
targets_dir <- file.path(tempdir(), "targets_test")
dir.create(targets_dir, showWarnings = FALSE, recursive = TRUE)
old_wd <- setwd(targets_dir)

info <- sw_pipeline_to_targets(pip_targets, input = 12)
stopifnot(file.exists(info$script))
cat("Targets script written to:", info$script, "\n")

# Now execute via the targets backend
tgt_result <- sw_pipeline_run_targets(
  pip_targets, input = 12,
  callr_function = NULL,
  trace = FALSE
)

stopifnot(is.list(tgt_result))
stopifnot("result" %in% names(tgt_result))
stopifnot("intermediates" %in% names(tgt_result))
stopifnot("steps_completed" %in% names(tgt_result))

# sqrt(12 * 3 + 5) = sqrt(41)
expected <- sqrt(12 * 3 + 5)
stopifnot(abs(tgt_result$result - expected) < 1e-10)
stopifnot(tgt_result$steps_completed == 3)
cat("targets result:", tgt_result$result, "(expected", expected, ")\n")

# Verify intermediates match sequential run
seq_result <- sw_pipeline_run(pip_targets, input = 12, trace = FALSE)
stopifnot(abs(tgt_result$result - seq_result$result) < 1e-10)
cat("Sequential and targets backends produce identical results.\n")

setwd(old_wd)
unlink(targets_dir, recursive = TRUE)
```

## Not covered in this tutorial

The following modules require external dependencies or API credentials
and are covered in their own dedicated vignettes:

- **Raw spectral unmixing** (`sw_autospectral_setup`, `sw_prepare_controls`,
  `sw_extract_af_spectra`, `sw_extract_spectral_variants`, `sw_unmix`,
  `sw_unmix_pipeline`) — requires AutoSpectral + real spectral control
  files. See the [Spectral Unmixing vignette](spectral-unmixing.qmd).

- **LLM assistant** (`sw_assistant`, `sw_assistant_configure`,
  `sw_assistant_batch_correction`, `sw_assistant_gating`,
  `sw_assistant_unmixing`, `sw_generate_pipeline_code`,
  `sw_quick_pipeline`, `sw_mcp_server`, `sw_pipeline_config_type`) —
  requires API credentials. See the [LLM Assistant vignette](llm-assistant.qmd).

## Session info

```{r}
#| label: session-info
cat("Tutorial Part 5 complete. All functions verified.\n\n")
sessionInfo()
```

## Functions covered

| Function | Module |
|----------|--------|
| `run_pipeline()` | pipeline |
| `sw_step()` | composable_pipeline |
| `execute_step()` | composable_pipeline |
| `sw_pipeline()` | composable_pipeline |
| `sw_pipeline_length()` | composable_pipeline |
| `sw_pipeline_step_names()` | composable_pipeline |
| `sw_pipeline_show()` | composable_pipeline |
| `sw_plot_pipeline()` | composable_pipeline |
| `sw_pipeline_add()` | composable_pipeline |
| `sw_pipeline_remove()` | composable_pipeline |
| `sw_pipeline_replace()` | composable_pipeline |
| `sw_pipeline_concat()` | composable_pipeline |
| `sw_pipeline_run()` | composable_pipeline |
| `%>>%` | composable_pipeline |
| `sw_step_read_fcs()` | composable_pipeline |
| `sw_step_remove_margins()` | composable_pipeline |
| `sw_step_signal_qc()` | composable_pipeline |
| `sw_step_normalize()` | composable_pipeline |
| `sw_step_create_som()` | composable_pipeline |
| `sw_step_correct_data()` | composable_pipeline |
| `sw_step_batch_correct()` | composable_pipeline |
| `sw_step_cluster()` | composable_pipeline |
| `sw_pipeline_to_targets()` | targets_backend |
| `sw_pipeline_run_targets()` | targets_backend |

**Previous:** [Part 4 — Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
