---
title: "Tutorial Part 1: Data Loading & Preprocessing"
vignette: >
  %\VignetteIndexEntry{Tutorial Part 1: Data Loading & Preprocessing}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This is **Part 1** of the SpectraWeaveR end-to-end tutorial series:

1. **Data Loading & Preprocessing** (this vignette)
2. [Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)
3. [Batch Correction](tutorial-03-batch-correction.qmd)
4. [Clustering & Visualization](tutorial-04-clustering-visualization.qmd)
5. [Pipeline Frameworks](tutorial-05-pipeline-frameworks.qmd)

This part covers: creating synthetic FCS files, reading them with
Aurora-safe defaults, inspecting channels, renaming markers, converting
between data formats, removing margin events and doublets, and tracking
events through preprocessing.

## Create synthetic FCS data

We generate 6 synthetic FCS files (3 samples per batch, 2 batches) with
10 channels: 4 scatter channels and 6 fluorescence detectors. Three
Gaussian sub-populations (CD4+ T cells, CD8+ T cells, B cells) give
clustering something meaningful to find later.

```{r}
#| label: synthetic-data
library(SpectraWeaveR)
library(flowCore)

set.seed(42)

# --- Shared helper used across all tutorial vignettes ---
make_tutorial_fcs <- function(fcs_dir, n_per_pop = 150, batch_shift = 0) {
  markers <- c("BV421-A", "BV510-A", "BV605-A", "BV711-A", "PE-A", "APC-A")
  n <- n_per_pop * 3

  # Three sub-populations with distinct marker profiles
  pop_cd4  <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 150000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 120000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  50000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  40000,  9000)),
    `BV421-A` = abs(rnorm(n_per_pop, 3000 + batch_shift, 500)),
    `BV510-A` = abs(rnorm(n_per_pop, 5000 + batch_shift, 800)),
    `BV605-A` = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    `BV711-A` = abs(rnorm(n_per_pop,  300 + batch_shift, 100)),
    `PE-A`    = abs(rnorm(n_per_pop,  200 + batch_shift, 100)),
    `APC-A`   = abs(rnorm(n_per_pop,  500 + batch_shift, 200))
  )

  pop_cd8  <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 140000, 20000)),
    `FSC-H` = abs(rnorm(n_per_pop, 110000, 18000)),
    `SSC-A` = abs(rnorm(n_per_pop,  45000, 10000)),
    `SSC-H` = abs(rnorm(n_per_pop,  36000,  9000)),
    `BV421-A` = abs(rnorm(n_per_pop, 2500 + batch_shift, 500)),
    `BV510-A` = abs(rnorm(n_per_pop,  400 + batch_shift, 150)),
    `BV605-A` = abs(rnorm(n_per_pop, 4500 + batch_shift, 700)),
    `BV711-A` = abs(rnorm(n_per_pop,  350 + batch_shift, 120)),
    `PE-A`    = abs(rnorm(n_per_pop,  250 + batch_shift, 100)),
    `APC-A`   = abs(rnorm(n_per_pop,  450 + batch_shift, 180))
  )

  pop_bcell <- cbind(
    `FSC-A` = abs(rnorm(n_per_pop, 130000, 18000)),
    `FSC-H` = abs(rnorm(n_per_pop, 100000, 16000)),
    `SSC-A` = abs(rnorm(n_per_pop,  40000,  8000)),
    `SSC-H` = abs(rnorm(n_per_pop,  32000,  7000)),
    `BV421-A` = abs(rnorm(n_per_pop,  300 + batch_shift, 120)),
    `BV510-A` = abs(rnorm(n_per_pop,  350 + batch_shift, 130)),
    `BV605-A` = abs(rnorm(n_per_pop,  280 + batch_shift, 100)),
    `BV711-A` = abs(rnorm(n_per_pop, 4000 + batch_shift, 600)),
    `PE-A`    = abs(rnorm(n_per_pop, 3500 + batch_shift, 500)),
    `APC-A`   = abs(rnorm(n_per_pop, 2000 + batch_shift, 400))
  )

  mat <- rbind(pop_cd4, pop_cd8, pop_bcell)
  mat <- mat[sample(nrow(mat)), ]
  storage.mode(mat) <- "double"
  ff <- flowFrame(mat)
  ff
}

# Create temp directory and write 6 FCS files
fcs_dir <- file.path(tempdir(), "tutorial_fcs")
dir.create(fcs_dir, showWarnings = FALSE, recursive = TRUE)

sample_info <- data.frame(
  file   = paste0("sample", 1:6, ".fcs"),
  sample = paste0("S", 1:6),
  batch  = rep(c("B1", "B2"), each = 3),
  condition = rep(c("control", "treatment"), 3),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(sample_info))) {
  shift <- if (sample_info$batch[i] == "B2") 300 else 0
  ff <- make_tutorial_fcs(fcs_dir, n_per_pop = 150, batch_shift = shift)
  write.FCS(ff, file.path(fcs_dir, sample_info$file[i]))
}

sample_meta <- tibble::as_tibble(sample_info)
cat("Wrote", nrow(sample_meta), "FCS files to", fcs_dir, "\n")
print(sample_meta)
```

## Reading FCS files

`sw_io_read_fcs()` wraps `flowCore::read.FCS()` with `truncate_max_range = FALSE`,
which is critical for Aurora data where fluorescence intensities can reach
~4 million and would otherwise be clipped.

```{r}
#| label: read-fcs
# Single file → flowFrame
fcs_path <- file.path(fcs_dir, "sample1.fcs")
ff_single <- sw_io_read_fcs(fcs_path)
stopifnot(is(ff_single, "flowFrame"))
cat("Single file:", nrow(ff_single), "events,", ncol(ff_single), "channels\n")

# Multiple files → flowSet
all_paths <- file.path(fcs_dir, sample_meta$file)
fs <- sw_io_read_fcs(all_paths)
stopifnot(is(fs, "flowSet"))
cat("flowSet:", length(fs), "samples\n")
```

`sw_io_load_unmixed()` is a directory-level loader that reads all FCS files
matching a pattern:

```{r}
#| label: load-unmixed
fs_loaded <- sw_io_load_unmixed(fcs_dir)
stopifnot(is(fs_loaded, "flowSet"))
stopifnot(length(fs_loaded) == 6)
cat("Loaded", length(fs_loaded), "samples via sw_io_load_unmixed()\n")
```

## Channel utilities

### Identify fluorescence channels

```{r}
#| label: fluor-channels
fluor_ch <- sw_channel_get_fluor(ff_single)
cat("Fluorescence channels:", paste(fluor_ch, collapse = ", "), "\n")
stopifnot(length(fluor_ch) == 6)
stopifnot(!any(grepl("^(FSC|SSC)", fluor_ch)))
```

### Classify signal vs fluorescence columns

`sw_channel_is_signal()` returns `TRUE` for all real signal channels
(scatter + fluorescence), excluding metadata like Time and Original_ID.
`sw_channel_is_fluor()` further excludes scatter channels.

```{r}
#| label: classify-cols
signal_mask <- sw_channel_is_signal(ff_single)
fluor_mask  <- sw_channel_is_fluor(ff_single)

cat("Signal columns:\n")
print(signal_mask)
cat("\nFluor columns:\n")
print(fluor_mask)

# All channels are signal (no Time column in synthetic data)
stopifnot(all(signal_mask))
# Scatter channels are signal but not fluor
stopifnot(fluor_mask["BV421-A"] == TRUE)
stopifnot(fluor_mask["FSC-A"] == FALSE)
```

### Rename detector channels to marker names

In real Aurora workflows, channels have detector names (BV421-A, PE-A, etc.)
and you map them to biological marker names (CD3, CD4, etc.) via
`sw_channel_set_markers()`.

```{r}
#| label: set-marker-names
marker_map <- c(
  "BV421-A" = "CD3",  "BV510-A" = "CD4",  "BV605-A" = "CD8",
  "BV711-A" = "CD19", "PE-A"    = "CD56", "APC-A"   = "HLADR"
)

ff_named <- sw_channel_set_markers(ff_single, marker_map)

pdata <- pData(parameters(ff_named))
named_markers <- pdata$desc[!is.na(pdata$desc)]
cat("Marker names set:", paste(named_markers, collapse = ", "), "\n")
stopifnot(all(c("CD3", "CD4", "CD8", "CD19", "CD56", "HLADR") %in% named_markers))
```

## Format conversions

SpectraWeaveR provides three conversion functions that bridge the gap
between flowCore objects and tibbles used by cyCombine / dplyr.

### flowFrame → tibble

```{r}
#| label: ff-to-tibble
tbl <- sw_io_ff_to_tibble(ff_single, sample = "S1", batch = "B1",
                               condition = "control")
stopifnot(is.data.frame(tbl))
stopifnot(nrow(tbl) == nrow(ff_single))
stopifnot(all(c("BV421-A", "FSC-A", "sample", "batch", "condition") %in% names(tbl)))
cat("Converted flowFrame to tibble:", nrow(tbl), "rows,", ncol(tbl), "cols\n")
head(tbl, 3)
```

### Expression matrix → tibble

```{r}
#| label: exprs-to-tibble
mat <- exprs(ff_single)
tbl_from_mat <- sw_io_exprs_to_tibble(mat, sample = "S1", batch = "B1")
stopifnot(is.data.frame(tbl_from_mat))
stopifnot(nrow(tbl_from_mat) == nrow(mat))
stopifnot("sample" %in% names(tbl_from_mat))
cat("Converted matrix to tibble:", nrow(tbl_from_mat), "rows\n")
```

### tibble → flowFrame (round-trip)

```{r}
#| label: tibble-to-ff
marker_cols <- c("BV421-A", "BV510-A", "BV605-A", "BV711-A", "PE-A", "APC-A")
ff_back <- sw_io_tibble_to_ff(tbl, markers = marker_cols)
stopifnot(is(ff_back, "flowFrame"))
stopifnot(nrow(ff_back) == nrow(tbl))
stopifnot(all(marker_cols %in% colnames(ff_back)))

# Verify round-trip fidelity
orig_vals <- exprs(ff_single)[, marker_cols]
back_vals <- exprs(ff_back)
stopifnot(all.equal(orig_vals, back_vals, tolerance = 1e-10))
cat("Round-trip flowFrame -> tibble -> flowFrame: values match\n")
```

## Margin removal

Margin events are cells that saturate detector limits. They must be
removed **before** any data transformation because `RemoveMargins` relies
on raw signal intensities.

### Standard margin removal

```{r}
#| label: remove-margins
ff_clean <- sw_filter_margins(ff_single)
stopifnot(is(ff_clean, "flowFrame"))
stopifnot(nrow(ff_clean) <= nrow(ff_single))
cat("Margin removal:", nrow(ff_single), "->", nrow(ff_clean), "events\n")
```

### PeacoQC-based margin removal

`sw_filter_margins_peacoqc()` is an enhanced wrapper that auto-selects
signal channels and supports per-channel range specifications.

```{r}
#| label: remove-margins-peacoqc
ff_clean_pqc <- sw_filter_margins_peacoqc(ff_single)
stopifnot(is(ff_clean_pqc, "flowFrame"))
stopifnot(nrow(ff_clean_pqc) <= nrow(ff_single))
cat("PeacoQC margin removal:", nrow(ff_single), "->",
    nrow(ff_clean_pqc), "events\n")
```

## Doublet removal

Doublets are two cells stuck together that appear as a single event with
anomalous area-to-height ratios.

### Singlets gate (parallelogram)

`sw_gate_singlets()` creates a parallelogram polygon gate from
FSC-A vs FSC-H. You can then apply it with `flowCore::filter()`.

```{r}
#| label: singlets-gate
gate <- sw_gate_singlets(ff_clean, channel1 = "FSC-A", channel2 = "FSC-H")
stopifnot(is(gate, "polygonGate"))
cat("Created singlets gate:", gate@filterId, "\n")

# Apply the gate
filt_result <- filter(ff_clean, gate)
ff_singlets <- Subset(ff_clean, filt_result)
stopifnot(is(ff_singlets, "flowFrame"))
cat("After singlets gate:", nrow(ff_clean), "->", nrow(ff_singlets), "events\n")
```

### Remove doublets (CytoPipeline algorithm)

`sw_filter_doublets()` applies singlet gates on one or two channel pairs
(FSC-A/FSC-H and optionally SSC-A/SSC-H) and returns the filtered flowFrame.

```{r}
#| label: remove-doublets
ff_no_doublets <- sw_filter_doublets(
  ff_clean,
  area_channels   = c("FSC-A"),
  height_channels = c("FSC-H"),
  nmads = 4
)
stopifnot(is(ff_no_doublets, "flowFrame"))
stopifnot(nrow(ff_no_doublets) <= nrow(ff_clean))
cat("Doublet removal:", nrow(ff_clean), "->", nrow(ff_no_doublets), "events\n")
```

### Remove doublets (PeacoQC algorithm)

```{r}
#| label: remove-doublets-peacoqc
ff_no_doublets_pqc <- sw_filter_doublets_peacoqc(
  ff_clean,
  area_channels   = c("FSC-A"),
  height_channels = c("FSC-H"),
  nmads = 4
)
stopifnot(is(ff_no_doublets_pqc, "flowFrame"))
stopifnot(nrow(ff_no_doublets_pqc) <= nrow(ff_clean))
cat("PeacoQC doublet removal:", nrow(ff_clean), "->",
    nrow(ff_no_doublets_pqc), "events\n")
```

## Debris removal

`sw_filter_debris()` applies a manual polygon gate in FSC-A vs SSC-A
space. You provide the polygon vertices as a numeric vector (first half =
x-coords, second half = y-coords).

```{r}
#| label: remove-debris
# Define a generous polygon that keeps most cells
gate_coords <- c(
  80000, 80000, 200000, 200000,    # FSC-A x-coords (4 vertices)
  10000, 80000,  80000,  10000     # SSC-A y-coords (4 vertices)
)

ff_no_debris <- sw_filter_debris(ff_clean, gate_data = gate_coords)
stopifnot(is(ff_no_debris, "flowFrame"))
stopifnot(nrow(ff_no_debris) <= nrow(ff_clean))
cat("Debris removal:", nrow(ff_clean), "->", nrow(ff_no_debris), "events\n")
```

## Aggregate and subsample

`sw_io_subsample()` pools events from multiple samples in a
flowSet and subsamples to a target total. Useful for creating a combined
overview before analysis.

```{r}
#| label: aggregate
ff_agg <- sw_io_subsample(fs, n_total_events = 500, seed = 42)
stopifnot(is(ff_agg, "flowFrame"))
stopifnot(nrow(ff_agg) <= 500)
stopifnot("File" %in% colnames(ff_agg))
stopifnot("Original_ID" %in% colnames(ff_agg))
cat("Aggregated", length(fs), "samples into", nrow(ff_agg), "events\n")
```

## Event audit trail

`sw_io_event_audit()` tracks how many events survive each
preprocessing step, providing a full audit trail.

```{r}
#| label: events-retained
intermediates <- list(
  raw           = nrow(ff_single),
  after_margins = nrow(ff_clean),
  after_singlets = nrow(ff_no_doublets),
  after_debris  = nrow(ff_no_debris)
)

audit <- sw_io_event_audit(intermediates)
stopifnot(is.data.frame(audit))
stopifnot(all(c("step", "n_events", "pct_of_initial", "pct_of_previous") %in%
              names(audit)))
print(audit)
```

## Clean up

```{r}
#| label: cleanup
unlink(fcs_dir, recursive = TRUE)
cat("Tutorial Part 1 complete. All functions verified.\n")
```

## Functions covered

| Function | Module |
|----------|--------|
| `sw_io_read_fcs()` | utils |
| `sw_io_load_unmixed()` | unmix |
| `sw_channel_get_fluor()` | utils |
| `sw_channel_is_signal()` | utils |
| `sw_channel_is_fluor()` | utils |
| `sw_channel_set_markers()` | utils |
| `sw_io_ff_to_tibble()` | utils |
| `sw_io_exprs_to_tibble()` | utils |
| `sw_io_tibble_to_ff()` | utils |
| `sw_filter_margins()` | unmix |
| `sw_filter_margins_peacoqc()` | gating_utils |
| `sw_gate_singlets()` | gating_utils |
| `sw_filter_doublets()` | gating_utils |
| `sw_filter_doublets_peacoqc()` | gating_utils |
| `sw_filter_debris()` | gating_utils |
| `sw_io_subsample()` | utils |
| `sw_io_event_audit()` | utils |

**Next:** [Part 2 — Gating & Signal QC](tutorial-02-gating-signal-qc.qmd)
